---
title: "Using GCSC"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

We are going to use GCSC to test for enriched gene sets/properties in ALS TWAS results.

***

# Run using provided coregulation matrix

***

## Download GCSC coregulation scores

```{bash, echo=T, eval=F}

cd /users/k1806347/oliverpainfel/Data/GCSC

file=$(echo Adipose_Subcutaneous_coregscores.npz Adipose_Subcutaneous_geneNames.txt Adipose_Visceral_Omentum_coregscores.npz Adipose_Visceral_Omentum_geneNames.txt Adrenal_Gland_coregscores.npz Adrenal_Gland_geneNames.txt Artery_Aorta_coregscores.npz Artery_Aorta_geneNames.txt Artery_Coronary_coregscores.npz Artery_Coronary_geneNames.txt Artery_Tibial_coregscores.npz Artery_Tibial_geneNames.txt Brain_Amygdala_coregscores.npz Brain_Amygdala_geneNames.txt Brain_Anterior_cingulate_cortex_BA24_coregscores.npz Brain_Anterior_cingulate_cortex_BA24_geneNames.txt Brain_Caudate_basal_ganglia_coregscores.npz Brain_Caudate_basal_ganglia_geneNames.txt Brain_Cerebellar_Hemisphere_coregscores.npz Brain_Cerebellar_Hemisphere_geneNames.txt Brain_Cerebellum_coregscores.npz Brain_Cerebellum_geneNames.txt Brain_Cortex_coregscores.npz Brain_Cortex_geneNames.txt Brain_Frontal_Cortex_BA9_coregscores.npz Brain_Frontal_Cortex_BA9_geneNames.txt Brain_Hippocampus_coregscores.npz Brain_Hippocampus_geneNames.txt Brain_Hypothalamus_coregscores.npz Brain_Hypothalamus_geneNames.txt Brain_Nucleus_accumbens_basal_ganglia_coregscores.npz Brain_Nucleus_accumbens_basal_ganglia_geneNames.txt Brain_Putamen_basal_ganglia_coregscores.npz Brain_Putamen_basal_ganglia_geneNames.txt Brain_Spinal_cord_cervical_c-1_coregscores.npz Brain_Spinal_cord_cervical_c-1_geneNames.txt Brain_Substantia_nigra_coregscores.npz Brain_Substantia_nigra_geneNames.txt Breast_Mammary_Tissue_coregscores.npz Breast_Mammary_Tissue_geneNames.txt Cells_EBV-transformed_lymphocytes_coregscores.npz Cells_EBV-transformed_lymphocytes_geneNames.txt Cells_Transformed_fibroblasts_coregscores.npz Cells_Transformed_fibroblasts_geneNames.txt Colon_Sigmoid_coregscores.npz Colon_Sigmoid_geneNames.txt Colon_Transverse_coregscores.npz Colon_Transverse_geneNames.txt Esophagus_Gastroesophageal_Junction_coregscores.npz Esophagus_Gastroesophageal_Junction_geneNames.txt Esophagus_Mucosa_coregscores.npz Esophagus_Mucosa_geneNames.txt Esophagus_Muscularis_coregscores.npz Esophagus_Muscularis_geneNames.txt Heart_Atrial_Appendage_coregscores.npz Heart_Atrial_Appendage_geneNames.txt Heart_Left_Ventricle_coregscores.npz Heart_Left_Ventricle_geneNames.txt Liver_coregscores.npz Liver_geneNames.txt Lung_coregscores.npz Lung_geneNames.txt Minor_Salivary_Gland_coregscores.npz Minor_Salivary_Gland_geneNames.txt Muscle_Skeletal_coregscores.npz Muscle_Skeletal_geneNames.txt Nerve_Tibial_coregscores.npz Nerve_Tibial_geneNames.txt Ovary_coregscores.npz Ovary_geneNames.txt Pancreas_coregscores.npz Pancreas_geneNames.txt Pituitary_coregscores.npz Pituitary_geneNames.txt Prostate_coregscores.npz Prostate_geneNames.txt Skin_Not_Sun_Exposed_Suprapubic_coregscores.npz Skin_Not_Sun_Exposed_Suprapubic_geneNames.txt Skin_Sun_Exposed_Lower_leg_coregscores.npz Skin_Sun_Exposed_Lower_leg_geneNames.txt Small_Intestine_Terminal_Ileum_coregscores.npz Small_Intestine_Terminal_Ileum_geneNames.txt Spleen_coregscores.npz Spleen_geneNames.txt Stomach_coregscores.npz Stomach_geneNames.txt Testis_coregscores.npz Testis_geneNames.txt Thyroid_coregscores.npz Thyroid_geneNames.txt Uterus_coregscores.npz Uterus_geneNames.txt Vagina_coregscores.npz Vagina_geneNames.txt Whole_Blood_coregscores.npz Whole_Blood_geneNames.txt)

for file_i in ${file}; do
wget https://storage.googleapis.com/broad-alkesgroup-public/GCSC/Coreg_scores/${file_i}
done

```

## Download GTEx v7 TWAS weights

```{bash, echo=T, eval=F}

cd /users/k1806347/oliverpainfel/Data/FUSION

file=$(echo GTEx.Adipose_Subcutaneous.P01.tar.bz2 GTEx.Adipose_Visceral_Omentum.P01.tar.bz2 GTEx.Adrenal_Gland.P01.tar.bz2 GTEx.Artery_Aorta.P01.tar.bz2 GTEx.Artery_Coronary.P01.tar.bz2 GTEx.Artery_Tibial.P01.tar.bz2 GTEx.Brain_Amygdala.P01.tar.bz2 GTEx.Brain_Anterior_cingulate_cortex_BA24.P01.tar.bz2 GTEx.Brain_Caudate_basal_ganglia.P01.tar.bz2 GTEx.Brain_Cerebellar_Hemisphere.P01.tar.bz2 GTEx.Brain_Cerebellum.P01.tar.bz2 GTEx.Brain_Cortex.P01.tar.bz2 GTEx.Brain_Frontal_Cortex_BA9.P01.tar.bz2 GTEx.Brain_Hippocampus.P01.tar.bz2 GTEx.Brain_Hypothalamus.P01.tar.bz2 GTEx.Brain_Nucleus_accumbens_basal_ganglia.P01.tar.bz2 GTEx.Brain_Putamen_basal_ganglia.P01.tar.bz2 GTEx.Brain_Spinal_cord_cervical_c-1.P01.tar.bz2 GTEx.Brain_Substantia_nigra.P01.tar.bz2 GTEx.Breast_Mammary_Tissue.P01.tar.bz2 GTEx.Cells_EBV-transformed_lymphocytes.P01.tar.bz2 GTEx.Cells_Transformed_fibroblasts.P01.tar.bz2 GTEx.Colon_Sigmoid.P01.tar.bz2 GTEx.Colon_Transverse.P01.tar.bz2 GTEx.Esophagus_Gastroesophageal_Junction.P01.tar.bz2 GTEx.Esophagus_Mucosa.P01.tar.bz2 GTEx.Esophagus_Muscularis.P01.tar.bz2 GTEx.Heart_Atrial_Appendage.P01.tar.bz2 GTEx.Heart_Left_Ventricle.P01.tar.bz2 GTEx.Liver.P01.tar.bz2 GTEx.Lung.P01.tar.bz2 GTEx.Minor_Salivary_Gland.P01.tar.bz2 GTEx.Muscle_Skeletal.P01.tar.bz2 GTEx.Nerve_Tibial.P01.tar.bz2 GTEx.Ovary.P01.tar.bz2 GTEx.Pancreas.P01.tar.bz2 GTEx.Pituitary.P01.tar.bz2 GTEx.Prostate.P01.tar.bz2 GTEx.Skin_Not_Sun_Exposed_Suprapubic.P01.tar.bz2 GTEx.Skin_Sun_Exposed_Lower_leg.P01.tar.bz2 GTEx.Small_Intestine_Terminal_Ileum.P01.tar.bz2 GTEx.Spleen.P01.tar.bz2 GTEx.Stomach.P01.tar.bz2 GTEx.Testis.P01.tar.bz2 GTEx.Thyroid.P01.tar.bz2 GTEx.Uterus.P01.tar.bz2 GTEx.Vagina.P01.tar.bz2 GTEx.Whole_Blood.P01.tar.bz2)

for file_i in ${file}; do
wget http://gusevlab.org/projects/fusion/weights/${file_i}
done

for file_i in $(ls *.tar.bz2); do
tar xjvf ${file_i}
done

rm *.tar.bz2

```

***

## Run TWAS using GTEx v7 SNP weights

```{bash, echo=T, eval=F}

gwas=ALS_only

weight=$(echo Adipose_Subcutaneous Adipose_Visceral_Omentum Adrenal_Gland Artery_Aorta Artery_Coronary Artery_Tibial Brain_Amygdala Brain_Anterior_cingulate_cortex_BA24 Brain_Caudate_basal_ganglia Brain_Cerebellar_Hemisphere Brain_Cerebellum Brain_Cortex Brain_Frontal_Cortex_BA9 Brain_Hippocampus Brain_Hypothalamus Brain_Nucleus_accumbens_basal_ganglia Brain_Putamen_basal_ganglia Brain_Spinal_cord_cervical_c-1 Brain_Substantia_nigra Breast_Mammary_Tissue Cells_EBV-transformed_lymphocytes Cells_Transformed_fibroblasts Colon_Sigmoid Colon_Transverse Esophagus_Gastroesophageal_Junction Esophagus_Mucosa Esophagus_Muscularis Heart_Atrial_Appendage Heart_Left_Ventricle Liver Lung Minor_Salivary_Gland Muscle_Skeletal Nerve_Tibial Ovary Pancreas Pituitary Prostate Skin_Not_Sun_Exposed_Suprapubic Skin_Sun_Exposed_Lower_leg Small_Intestine_Terminal_Ileum Spleen Stomach Testis Thyroid Uterus Vagina Whole_Blood)

chr=$(seq 1 22)

for gwas_i in ${gwas}; do
  mkdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only
  for weight_i in ${weight}; do
    mkdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only/${weight_i}
    for chr_i in ${chr}; do
  
      sbatch -p cpu --mem 10G -n 1 /users/k1806347/oliverpainfel/Software/Rscript.sh /users/k1806347/oliverpainfel/biomarkers-brc-mh/TWAS_resource/FUSION/fusion_twas-master/FUSION.assoc_test.R \
          --sumstats /users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/resources/data/gwas_sumstat/ALS_only/ALS_only.cleaned.munged.sumstats.gz \
          --weights /users/k1806347/oliverpainfel/Data/FUSION/${weight_i}.P01.pos \
          --weights_dir /users/k1806347/oliverpainfel/Data/FUSION/ \
          --ref_ld_chr /users/k1806347/oliverpainfel/biomarkers-brc-mh/TWAS_resource/FUSION/LDREF/1000G.EUR. \
          --out /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only/${weight_i}/ALS_only_${weight_i}_chr${chr_i}.dat \
          --chr ${chr_i}
          
    done
  done
done

```

***

## Run GCSC using example gene sets

```{bash, echo=T, eval=F}

##
# All tissues
##

mkdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_allTissues_test

N=$(cat /users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/resources/data/gwas_sumstat/ALS_only/ALS_only.cleaned.munged.median_N.txt)

weight=$(echo Adipose_Subcutaneous Adipose_Visceral_Omentum Adrenal_Gland Artery_Aorta Artery_Coronary Artery_Tibial Brain_Amygdala Brain_Anterior_cingulate_cortex_BA24 Brain_Caudate_basal_ganglia Brain_Cerebellar_Hemisphere Brain_Cerebellum Brain_Cortex Brain_Frontal_Cortex_BA9 Brain_Hippocampus Brain_Hypothalamus Brain_Nucleus_accumbens_basal_ganglia Brain_Putamen_basal_ganglia Brain_Spinal_cord_cervical_c-1 Brain_Substantia_nigra Breast_Mammary_Tissue Cells_EBV-transformed_lymphocytes Cells_Transformed_fibroblasts Colon_Sigmoid Colon_Transverse Esophagus_Gastroesophageal_Junction Esophagus_Mucosa Esophagus_Muscularis Heart_Atrial_Appendage Heart_Left_Ventricle Liver Lung Minor_Salivary_Gland Muscle_Skeletal Nerve_Tibial Ovary Pancreas Pituitary Prostate Skin_Not_Sun_Exposed_Suprapubic Skin_Sun_Exposed_Lower_leg Small_Intestine_Terminal_Ileum Spleen Stomach Testis Thyroid Uterus Vagina Whole_Blood)

python /users/k1806347/oliverpainfel/Software/GCSC/gcsc.py \
  --geneSets /users/k1806347/oliverpainfel/Data/GCSC/SampleGeneSets.csv \
  --TWASdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only/tissue \
  --N ${N} \
  --tissues ${weight} \
  --coreg /users/k1806347/oliverpainfel/Data/GCSC \
  --out /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_allTissues_test

##
# Brain and Blood tissues
##

mkdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_Brain_Blood_test

N=$(cat /users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/resources/data/gwas_sumstat/ALS_only/ALS_only.cleaned.munged.median_N.txt)

weight=$(echo Brain_Amygdala Brain_Anterior_cingulate_cortex_BA24 Brain_Caudate_basal_ganglia Brain_Cerebellar_Hemisphere Brain_Cerebellum Brain_Cortex Brain_Frontal_Cortex_BA9 Brain_Hippocampus Brain_Hypothalamus Brain_Nucleus_accumbens_basal_ganglia Brain_Putamen_basal_ganglia Brain_Spinal_cord_cervical_c-1 Brain_Substantia_nigra Whole_Blood)

python /users/k1806347/oliverpainfel/Software/GCSC/gcsc.py \
  --geneSets /users/k1806347/oliverpainfel/Data/GCSC/SampleGeneSets.csv \
  --TWASdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only/tissue \
  --N ${N} \
  --tissues ${weight} \
  --coreg /users/k1806347/oliverpainfel/Data/GCSC \
  --out /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_Brain_Blood_test

##
# Brain tissues
##

mkdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_Brain_test

N=$(cat /users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/resources/data/gwas_sumstat/ALS_only/ALS_only.cleaned.munged.median_N.txt)

weight=$(echo Brain_Amygdala Brain_Anterior_cingulate_cortex_BA24 Brain_Caudate_basal_ganglia Brain_Cerebellar_Hemisphere Brain_Cerebellum Brain_Cortex Brain_Frontal_Cortex_BA9 Brain_Hippocampus Brain_Hypothalamus Brain_Nucleus_accumbens_basal_ganglia Brain_Putamen_basal_ganglia Brain_Spinal_cord_cervical_c-1 Brain_Substantia_nigra)

python /users/k1806347/oliverpainfel/Software/GCSC/gcsc.py \
  --geneSets /users/k1806347/oliverpainfel/Data/GCSC/SampleGeneSets.csv \
  --TWASdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only/tissue \
  --N ${N} \
  --tissues ${weight} \
  --coreg /users/k1806347/oliverpainfel/Data/GCSC \
  --out /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_Brain_test


```

***

## Run GCSC using drug gene sets (binary and non binary)

***

### Prepare the gene-drug interaction data

```{r, echo=T, eval=F}
# Format the DrugTargetor data for GCSC

library(data.table)

# Read in database
pathways<-fread('/scratch/prj/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/resources/data/drug_targetor/wholedatabase_for_targetor')

# Read in GCSC gene universe
universe<-fread('/scratch/prj/oliverpainfel/Data/GCSC/gene_universe.txt', header=F)

gene_id<-data.frame(universe)
names(gene_id)<-'ID'

# Use BioMart to convert gene IDs
library(biomaRt)
ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl", GRCh=37)
Genes<-getBM(attributes=c('ensembl_gene_id','external_gene_name'), mart = ensembl)
Genes<-Genes[!duplicated(Genes$ensembl_gene_id),]

pathways_id<-merge(pathways, Genes, by.x='gene', by.y='external_gene_name')

drugs<-unique(pathways$atc)
for(i in 1:length(drugs)){
  drug_tmp<-pathways_id[pathways_id$atc == drugs[i],]
  
  gene_id[[drugs[i]]]<-ifelse(gene_id$ID %in% drug_tmp$ensembl_gene_id[drug_tmp$activity_type %in% c('DECREASED_EXPRESSION','NEGATIVE_RESPONSE','OPPOSITE_RESPONSE')], -1, ifelse(gene_id$ID %in% drug_tmp$ensembl_gene_id[drug_tmp$activity_type %in% c('INCREASED_EXPRESSION','POSITIVE_RESPONSE')], 1, 0))
  

}

# Remove sets with less than 2 genes present in each gene set and in the TWAS
weight=c('Adipose_Subcutaneous','Adipose_Visceral_Omentum','Adrenal_Gland','Artery_Aorta','Artery_Coronary','Artery_Tibial','Brain_Amygdala','Brain_Anterior_cingulate_cortex_BA24','Brain_Caudate_basal_ganglia','Brain_Cerebellar_Hemisphere','Brain_Cerebellum','Brain_Cortex','Brain_Frontal_Cortex_BA9','Brain_Hippocampus','Brain_Hypothalamus','Brain_Nucleus_accumbens_basal_ganglia','Brain_Putamen_basal_ganglia','Brain_Spinal_cord_cervical_c-1','Brain_Substantia_nigra','Breast_Mammary_Tissue','Cells_EBV-transformed_lymphocytes','Cells_Transformed_fibroblasts','Colon_Sigmoid','Colon_Transverse','Esophagus_Gastroesophageal_Junction','Esophagus_Mucosa','Esophagus_Muscularis','Heart_Atrial_Appendage','Heart_Left_Ventricle','Liver','Lung','Minor_Salivary_Gland','Muscle_Skeletal','Nerve_Tibial','Ovary','Pancreas','Pituitary','Prostate','Skin_Not_Sun_Exposed_Suprapubic','Skin_Sun_Exposed_Lower_leg','Small_Intestine_Terminal_Ileum','Spleen','Stomach','Testis','Thyroid','Uterus','Vagina','Whole_Blood')

als_twas<-NULL
for(weight_i in weight){
  for(chr_i in 1:22){
    als_twas<-rbind(
                als_twas,
                fread(paste0('/users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only/',weight_i,'/ALS_only_',weight_i,'_chr',chr_i,'.dat')))
  }
}
als_twas$ensembl_id<-gsub('\\..*','',gsub('.*ENSG','ENSG',als_twas$FILE))
als_twas<-als_twas[!is.na(als_twas$TWAS.Z),]

# All tissues
gene_id_all<-gene_id[gene_id$ID %in% als_twas$ensembl_id,]

gene_id_all_2<-gene_id[,c(1, which(colSums(abs(gene_id_all[,-1])) >= 2)+1)]

gene_id_all_mat<-matrix(as.matrix(gene_id_all_2[,-1]), nrow=nrow(gene_id_all_2), ncol=ncol(gene_id_all_2)-1, dimnames=list(gene_id_all_2$ID,gsub(' ' ,'',gsub('[[:punct:]]','_',names(gene_id_all_2)[-1]))))
gene_id_all_mat<-t(gene_id_all_mat)

gene_id_all_mat_nondir<-gene_id_all_mat
gene_id_all_mat_nondir[gene_id_all_mat_nondir == -1]<-1
gene_id_all_mat_dir<-gene_id_all_mat

chunks<-split(1:nrow(gene_id_all_mat_nondir), ceiling(seq_along(1:nrow(gene_id_all_mat_nondir))/10))
for(chunks_i in 1:length(chunks)){
  write.csv(gene_id_all_mat_nondir[chunks[[chunks_i]],], paste0('/users/k1806347/oliverpainfel/Data/GCSC/drugtargetor_nondir_allTissues_',chunks_i,'.csv'), quote=F)

}

# Brain and Blood
gene_id_brain_blood<-gene_id[gene_id$ID %in% als_twas$ensembl_id[als_twas$PANEL %in% c('Brain_Amygdala','Brain_Anterior_cingulate_cortex_BA24','Brain_Caudate_basal_ganglia','Brain_Cerebellar_Hemisphere','Brain_Cerebellum','Brain_Cortex','Brain_Frontal_Cortex_BA9','Brain_Hippocampus','Brain_Hypothalamus','Brain_Nucleus_accumbens_basal_ganglia','Brain_Putamen_basal_ganglia','Brain_Spinal_cord_cervical_c-1','Brain_Substantia_nigra','Whole_Blood')],]

gene_id_brain_blood_2<-gene_id[,c(1, which(colSums(abs(gene_id_brain_blood[,-1])) >= 2)+1)]

gene_id_brain_blood_mat<-matrix(as.matrix(gene_id_brain_blood_2[,-1]), nrow=nrow(gene_id_brain_blood_2), ncol=ncol(gene_id_brain_blood_2)-1, dimnames=list(gene_id_brain_blood_2$ID,gsub(' ' ,'',gsub('[[:punct:]]','_',names(gene_id_brain_blood_2)[-1]))))
gene_id_brain_blood_mat<-t(gene_id_brain_blood_mat)

gene_id_brain_blood_mat_nondir<-gene_id_brain_blood_mat
gene_id_brain_blood_mat_nondir[gene_id_brain_blood_mat_nondir == -1]<-1
gene_id_brain_blood_mat_dir<-gene_id_brain_blood_mat

chunks<-split(1:nrow(gene_id_brain_blood_mat_nondir), ceiling(seq_along(1:nrow(gene_id_brain_blood_mat_nondir))/10))
for(chunks_i in 1:length(chunks)){
  write.csv(gene_id_brain_blood_mat_nondir[chunks[[chunks_i]],], paste0('/users/k1806347/oliverpainfel/Data/GCSC/drugtargetor_nondir_Brain_Blood_',chunks_i,'.csv'), quote=F)

}

# Brain
gene_id_brain<-gene_id[gene_id$ID %in% als_twas$ensembl_id[als_twas$PANEL %in% c('Brain_Amygdala','Brain_Anterior_cingulate_cortex_BA24','Brain_Caudate_basal_ganglia','Brain_Cerebellar_Hemisphere','Brain_Cerebellum','Brain_Cortex','Brain_Frontal_Cortex_BA9','Brain_Hippocampus','Brain_Hypothalamus','Brain_Nucleus_accumbens_basal_ganglia','Brain_Putamen_basal_ganglia','Brain_Spinal_cord_cervical_c-1','Brain_Substantia_nigra')],]

gene_id_brain_2<-gene_id[,c(1, which(colSums(abs(gene_id_brain[,-1])) >= 2)+1)]

gene_id_brain_mat<-matrix(as.matrix(gene_id_brain_2[,-1]), nrow=nrow(gene_id_brain_2), ncol=ncol(gene_id_brain_2)-1, dimnames=list(gene_id_brain_2$ID,gsub(' ' ,'',gsub('[[:punct:]]','_',names(gene_id_brain_2)[-1]))))
gene_id_brain_mat<-t(gene_id_brain_mat)

gene_id_brain_mat_nondir<-gene_id_brain_mat
gene_id_brain_mat_nondir[gene_id_brain_mat_nondir == -1]<-1
gene_id_brain_mat_dir<-gene_id_brain_mat

chunks<-split(1:nrow(gene_id_brain_mat_nondir), ceiling(seq_along(1:nrow(gene_id_brain_mat_nondir))/10))
for(chunks_i in 1:length(chunks)){
  write.csv(gene_id_brain_mat_nondir[chunks[[chunks_i]],], paste0('/users/k1806347/oliverpainfel/Data/GCSC/drugtargetor_nondir_Brain_',chunks_i,'.csv'), quote=F)

}


```

***

### Non-directional

```{bash, echo=T, eval=F}

##
# All tissues
##

N=$(cat /users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/resources/data/gwas_sumstat/ALS_only/ALS_only.cleaned.munged.median_N.txt)

weight=$(echo Adipose_Subcutaneous Adipose_Visceral_Omentum Adrenal_Gland Artery_Aorta Artery_Coronary Artery_Tibial Brain_Amygdala Brain_Anterior_cingulate_cortex_BA24 Brain_Caudate_basal_ganglia Brain_Cerebellar_Hemisphere Brain_Cerebellum Brain_Cortex Brain_Frontal_Cortex_BA9 Brain_Hippocampus Brain_Hypothalamus Brain_Nucleus_accumbens_basal_ganglia Brain_Putamen_basal_ganglia Brain_Spinal_cord_cervical_c-1 Brain_Substantia_nigra Breast_Mammary_Tissue Cells_EBV-transformed_lymphocytes Cells_Transformed_fibroblasts Colon_Sigmoid Colon_Transverse Esophagus_Gastroesophageal_Junction Esophagus_Mucosa Esophagus_Muscularis Heart_Atrial_Appendage Heart_Left_Ventricle Liver Lung Minor_Salivary_Gland Muscle_Skeletal Nerve_Tibial Ovary Pancreas Pituitary Prostate Skin_Not_Sun_Exposed_Suprapubic Skin_Sun_Exposed_Lower_leg Small_Intestine_Terminal_Ileum Spleen Stomach Testis Thyroid Uterus Vagina Whole_Blood)

for chunk in $(seq 1 $(ls /users/k1806347/oliverpainfel/Data/GCSC/drugtargetor_nondir_allTissues_*.csv | wc -l)); do

mkdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_allTissues_drugtargetor_nondir_chunk_${chunk}

sbatch -p cpu --mem 10G -n 1 --wrap="python /users/k1806347/oliverpainfel/Software/GCSC/gcsc.py \
  --geneSets /users/k1806347/oliverpainfel/Data/GCSC/drugtargetor_nondir_allTissues_${chunk}.csv \
  --TWASdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only/tissue \
  --N ${N} \
  --tissues ${weight} \
  --coreg /users/k1806347/oliverpainfel/Data/GCSC \
  --out /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_allTissues_drugtargetor_nondir_chunk_${chunk}"

done

##
# Brain and Blood tissues
##

N=$(cat /users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/resources/data/gwas_sumstat/ALS_only/ALS_only.cleaned.munged.median_N.txt)

weight=$(echo Brain_Amygdala Brain_Anterior_cingulate_cortex_BA24 Brain_Caudate_basal_ganglia Brain_Cerebellar_Hemisphere Brain_Cerebellum Brain_Cortex Brain_Frontal_Cortex_BA9 Brain_Hippocampus Brain_Hypothalamus Brain_Nucleus_accumbens_basal_ganglia Brain_Putamen_basal_ganglia Brain_Spinal_cord_cervical_c-1 Brain_Substantia_nigra Whole_Blood)

for chunk in $(seq 1 $(ls /users/k1806347/oliverpainfel/Data/GCSC/drugtargetor_nondir_Brain_Blood_*.csv | wc -l)); do

mkdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_Brain_Blood_drugtargetor_nondir_chunk_${chunk}

sbatch -p cpu --mem 10G -n 1 --wrap="python /users/k1806347/oliverpainfel/Software/GCSC/gcsc.py \
  --geneSets /users/k1806347/oliverpainfel/Data/GCSC/drugtargetor_nondir_Brain_Blood_${chunk}.csv \
  --TWASdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only/tissue \
  --N ${N} \
  --tissues ${weight} \
  --coreg /users/k1806347/oliverpainfel/Data/GCSC \
  --out /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_Brain_Blood_drugtargetor_nondir_chunk_${chunk}"

done

##
# Brain tissues
##

N=$(cat /users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/resources/data/gwas_sumstat/ALS_only/ALS_only.cleaned.munged.median_N.txt)

weight=$(echo Brain_Amygdala Brain_Anterior_cingulate_cortex_BA24 Brain_Caudate_basal_ganglia Brain_Cerebellar_Hemisphere Brain_Cerebellum Brain_Cortex Brain_Frontal_Cortex_BA9 Brain_Hippocampus Brain_Hypothalamus Brain_Nucleus_accumbens_basal_ganglia Brain_Putamen_basal_ganglia Brain_Spinal_cord_cervical_c-1 Brain_Substantia_nigra)

for chunk in $(seq 1 $(ls /users/k1806347/oliverpainfel/Data/GCSC/drugtargetor_nondir_Brain_*.csv | grep -v Blood | wc -l)); do

mkdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_Brain_drugtargetor_nondir_chunk_${chunk}

sbatch -p cpu --mem 10G -n 1 --wrap="python /users/k1806347/oliverpainfel/Software/GCSC/gcsc.py \
  --geneSets /users/k1806347/oliverpainfel/Data/GCSC/drugtargetor_nondir_Brain_${chunk}.csv \
  --TWASdir /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only/tissue \
  --N ${N} \
  --tissues ${weight} \
  --coreg /users/k1806347/oliverpainfel/Data/GCSC \
  --out /users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_Brain_drugtargetor_nondir_chunk_${chunk}"

done

```

***

## Results

```{r, eval=F, echo =T}

library(data.table)

tissues<-c('allTissues','Brain_Blood','Brain')

all<-list()
for(tissues_i in tissues){
  tmp<-list.files(path='/users/k1806347/oliverpainfel/Analyses/GCSC', pattern=paste0('ALS_only_GCSC_',tissues_i,'_drugtargetor_nondir_chunk'))
  res<-NULL
  for(i in 1:length(tmp)){
    res<-rbind(res, fread(paste0('/users/k1806347/oliverpainfel/Analyses/GCSC/ALS_only_GCSC_',tissues_i,'_drugtargetor_nondir_chunk_',i,'/GCSCresults.txt'), fill=T, sep=' '))
  }
  res<-res[grepl('enrichment:',res$Parameter),]
  res<-res[order(res$`P-value`),]
  all[[tissues_i]]<-res
}

res_all<-Reduce(function(dtf1, dtf2) merge(dtf1, dtf2, by = "Parameter", all.x = TRUE), all)
names(res_all)<-c('Drug','Est_all','SE_all','P_all','Est_brain_blood','SE_brain_blood','P_brain_blood','Est_brain','SE_brain','P_brain')

cor(res_all[,c('Est_all','Est_brain_blood','Est_brain'),with=F], use='p')
res_all<-res_all[order(res_all$P_brain_blood),]
res_all[res_all$Est_brain_blood > 0,]

res_all$P_all.FDR<-p.adjust(res_all$P_all, method='fdr')
res_all$P_brain_blood.FDR<-p.adjust(res_all$P_brain_blood, method='fdr')
res_all$P_brain.FDR<-p.adjust(res_all$P_brain, method='fdr')

sum(res_all$P_all.FDR < 0.05 & res_all$Est_all > 0, na.rm=T)
sum(res_all$P_brain_blood.FDR < 0.05 & res_all$Est_brain_blood > 0, na.rm=T)
sum(res_all$P_brain.FDR < 0.05 & res_all$Est_brain > 0, na.rm=T)

res_all[res_all$P_all.FDR < 0.05 & res_all$Est_all > 0,]
res_all[res_all$P_brain_blood.FDR < 0.05 & res_all$Est_brain_blood > 0,]
res_all[res_all$P_brain.FDR < 0.05 & res_all$Est_brain > 0,]

# Results are quite different when using all tissues. I think we should stick to brain and blood for consistency.
res_gs<-all[[2]]
res_gs$Drug<-gsub('_CID.*','',gsub('.*NAME_','',res_gs$Parameter))
res_gs$Drug<-tolower(res_gs$Drug)

res_gs$ATC<-gsub('ATC_','',gsub('_NAME.*','',res_gs$Parameter))

res_gs<-res_gs[,c('Drug','Value','Standard_error','P-value','ATC'), with=F]
names(res_gs)<-c('Drug','Enrichment','SE','P','ATC')
res_gs$P.FDR<-p.adjust(res_gs$P, method='fdr')

write.csv(res_gs, '/users/k1806347/oliverpainfel/Analyses/GCSC/GCSC_Brain_Blood_ALS_only_results.csv', row.names=F, quote=F)


# Perform enrichment analysis of ATC codes
res_gs_enrich<-NULL
for(i in 1:nrow(res_gs)){
  res_gs_enrich<-rbind(res_gs_enrich, data.frame( ATC=unlist(strsplit(res_gs$ATC[i],'_')),
                                                  Enrichment=res_gs$Enrichment[i]))
}

res_gs_enrich$atc_cat<-substr(res_gs_enrich$ATC, 1, 4) 

# Test for enrichment for each ATC category
atc_enrich<-NULL
for(cat in unique(res_gs_enrich$atc_cat)){
  class_bin<-rep(0, nrow(res_gs_enrich))
  class_bin[res_gs_enrich$atc_cat == cat]<-1
  
  if(sum(class_bin == 1) > 5){
    
    wil_cox_res<-wilcox.test(rank(res_gs_enrich$Enrichment) ~ class_bin, conf.int =T, alternative='less')
    
    atc_enrich<-rbind(atc_enrich, data.frame(ATC=cat,
                                             Estimate=as.numeric(wil_cox_res$estimate),
                                             Class_Median=median(res_gs_enrich$Enrichment[class_bin == 1]),
                                             Non_Class_Median=median(res_gs_enrich$Enrichment[class_bin == 0]),
                                             P=wil_cox_res$p.value,
                                             N=sum(class_bin)))
  }
}

atc<-fread('/users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/resources/data/atc/atc_20220201.txt', sep='!')
names(atc)<-c('Code','Name')
atc$Name<-tolower(atc$Name)

atc_labels<-atc[nchar(atc$Code) == 4,]
atc_enrich<-merge(atc_enrich, atc_labels, by.x='ATC', by.y='Code')
atc_enrich<-atc_enrich[order(atc_enrich$P),]
atc_enrich$P.FDR<-p.adjust(atc_enrich$P, method='fdr')

write.csv(atc_enrich, paste0('/users/k1806347/oliverpainfel/Analyses/GCSC/GCSC_Brain_Blood_ALS_only_atc_res.csv'), row.names=F)

# Perform enrichment again on level 4 ATC codes for enriched ATC level 3 codes.
# Test for enrichment for each ATC category
res_gs_enrich$atc_cat_2<-substr(res_gs_enrich$ATC, 1, 5) 
atc_enrich_2<-NULL
for(cat in unique(res_gs_enrich$atc_cat_2)){
  class_bin<-rep(0, nrow(res_gs_enrich))
  class_bin[res_gs_enrich$atc_cat_2 == cat]<-1
  
  if(sum(class_bin == 1) > 1){
    
    wil_cox_res<-wilcox.test(rank(res_gs_enrich$Enrichment) ~ class_bin, conf.int =T, alternative='less')
    
    atc_enrich_2<-rbind(atc_enrich_2, data.frame(ATC=cat,
                                             Estimate=as.numeric(wil_cox_res$estimate),
                                             Class_Median=median(res_gs_enrich$Enrichment[class_bin == 1]),
                                             Non_Class_Median=median(res_gs_enrich$Enrichment[class_bin == 0]),
                                             P=wil_cox_res$p.value,
                                             N=sum(class_bin)))
  }
}

atc_labels_2<-atc[nchar(atc$Code) == 5,]
atc_enrich_2<-merge(atc_enrich_2, atc_labels_2, by.x='ATC', by.y='Code')
atc_enrich_2<-atc_enrich_2[order(atc_enrich_2$P),]

write.csv(atc_enrich_2, paste0('/users/k1806347/oliverpainfel/Analyses/GCSC/GCSC_Brain_Blood_ALS_only_atc_res_level4.csv'), row.names=F)

# There is only one C08C level 4 category with more than 1 drug present.
# Within antiepileptics, N03AF - carboxamide derivatives and N03AB - hydantoin derivatives are most enriched.
# We want to show results for all ATC groups that were significant in MAGMA, TWAS-GSEA and GCSC, that is why we ran the analysis across all ATC groups.

```

GCSC won't tell us the direction of effect of the drug on ALS risk as it doesn't consider the direction of TWAS association. When using directional gene-sets it will tell use whether ALS-associated genes are up or down regulated by the drug. We could charaterise associated sets, by showing the concordance/discordance of ALS-risk and drug action. This is similar to DrugTargetor. The enrichment analysis will have lower power by not considering whether the drug consistently reverse ALS-risk genes.

***

### Create a plot comparing results across MAGMA, GCSC and TWAS-GSEA

```{r, eval=F, echo=T}

#######
# Drug associations
#######
###
# MAGMA
###
# One sided p
magma_gs<-fread(paste0('/users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/results/ALS_only/magma/magma_drug_targetor.clean.csv'))
magma_gs<-magma_gs[order(magma_gs$P),]
magma_gs$P.FDR<-p.adjust(magma_gs$P, method='fdr')
magma_gs$NAME<-paste(toupper(substr(magma_gs$NAME, 1, 1)), substr(magma_gs$NAME, 2, nchar(magma_gs$NAME)), sep="")

magma_gs$Z<--qnorm(magma_gs$P)

magma_gs<-magma_gs[,c('NAME','Z','P','P.FDR'),with=F]
magma_gs$Method<-'MAGMA'
magma_gs$Panel<-'MAGMA'

###
# GCSC
###
# Two sided p
gcsc_gs<-fread(paste0('/users/k1806347/oliverpainfel/Analyses/GCSC/GCSC_Brain_Blood_ALS_only_results.csv'))
gcsc_gs$Drug<-paste(toupper(substr(gcsc_gs$Drug, 1, 1)), substr(gcsc_gs$Drug, 2, nchar(gcsc_gs$Drug)), sep="")

gcsc_gs$Z<--qnorm(gcsc_gs$P/2)
gcsc_gs$Z<-gcsc_gs$Z*sign(gcsc_gs$Enrichment)

gcsc_gs<-gcsc_gs[,c('Drug','Z','P','P.FDR'),with=F]
names(gcsc_gs)<-c('NAME','Z','P','P.FDR')
gcsc_gs$Method<-'GCSC'
gcsc_gs$Panel<-'Brain and Blood'

###
# TWAS-GSEA
###
# Two sided p
weights<-read.table(paste0('/users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/results/ALS_only/twas/list_of_weights.txt'))$V1
weights<-weights[!grepl('SPLIC',weights)]

gsea_drugtargetor_all<-NULL
for(i in weights){
  gsea_drugtargetor<-fread(paste0('/users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/results/ALS_only/twas/drugtargetor/twas_gsea_drugtargetor_',i,'.competitive.clean.csv'))
  gsea_drugtargetor$Panel<-i
  gsea_drugtargetor_all<-rbind(gsea_drugtargetor_all, gsea_drugtargetor)
}

gsea_drugtargetor_all$P.FDR<-p.adjust(gsea_drugtargetor_all$P, method='fdr')

gsea_drugtargetor_all<-gsea_drugtargetor_all

gsea_drugtargetor_all$NAME<-paste(toupper(substr(gsea_drugtargetor_all$NAME, 1, 1)), substr(gsea_drugtargetor_all$NAME, 2, nchar(gsea_drugtargetor_all$NAME)), sep="")
gsea_drugtargetor_all$NAME<-gsub('\\..*','',gsea_drugtargetor_all$NAME)
  
gsea_gs<-gsea_drugtargetor_all
gsea_gs$Method<-'TWAS-GSEA'
gsea_gs<-gsea_gs[,c('NAME','T','P','P.FDR','Method','Panel'),with=F]
names(gsea_gs)<-c('NAME','Z','P','P.FDR','Method','Panel')
# Flip Z so >0 indicates reversal of ALS risk.
gsea_gs$Z<--gsea_gs$Z

# Update labels
gsea_gs$Panel[gsea_gs$Panel == "Brain_Anterior_cingulate_cortex_BA24"]<-"Brain - Anterior cingulate cortex BA24 (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "psychencode"]<-"Brain - DLPFC (PsychENCODE)"
gsea_gs$Panel[gsea_gs$Panel == "Brain_Cerebellar_Hemisphere"]<-"Brain - Cerebellar hemisphere (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "NTR.BLOOD.RNAARR"]<-"Blood (NTR)"
gsea_gs$Panel[gsea_gs$Panel == "Brain_Cortex"]<-"Brain - Cortex (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "Brain_Caudate_basal_ganglia"]<-"Brain - Caudate basal ganglia (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "YFS.BLOOD.RNAARR"]<-"Blood (YFS)"
gsea_gs$Panel[gsea_gs$Panel == "Brain_Hippocampus"]<-"Brain - Hippocampus (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "Brain_Substantia_nigra"]<-"Brain - Substantia nigra (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "Brain_Nucleus_accumbens_basal_ganglia"]<-"Brain - Nucleus accumbens basal ganglia (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "Brain_Putamen_basal_ganglia"]<-"Brain - Putamen basal ganglia (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "Brain_Frontal_Cortex_BA9"]<-"Brain - Frontal cortex BA9 (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "Whole_Blood"]<-"Blood (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "CMC.BRAIN.RNASEQ"]<-"Brain - DLPFC (CMC)"
gsea_gs$Panel[gsea_gs$Panel == "Brain_Cerebellum"]<-"Brain - Cerebellum (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "Brain_Spinal_cord_cervical_c-1"]<-"Brain - Spinal cord vervical c-1 (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "Brain_Hypothalamus"]<-"Brain - Hypothalamus (GTEx)"
gsea_gs$Panel[gsea_gs$Panel == "kcl_brainbank_motor_cortex"]<-"Brain - Motor cortex (KCL Brainbank)"
gsea_gs$Panel[gsea_gs_atc$Panel == "Brain_Amygdala"]<-"Brain - Amygdala (GTEx)"

###
# Combine results
###

all_gs<-do.call(rbind, list(magma_gs, gcsc_gs, gsea_gs))
all_gs<-all_gs[all_gs$NAME %in% all_gs$NAME[all_gs$P.FDR < 0.05 & all_gs$Z > 0],]
all_gs$Method<-factor(all_gs$Method, levels=c('MAGMA','GCSC','TWAS-GSEA'))

library(ggplot2)
library(cowplot)

group_siz<-NULL
for(i in c('MAGMA','GCSC','TWAS-GSEA')){
  group_siz<-rbind(group_siz, data.frame(Group=i,
                           Size=length(unique(all_gs$Panel[all_gs$Method==i]))))
}

# Set minimum size to 3 to allow space for labels
group_siz$Size[group_siz$Size < 2]<-2
group_siz$Prop<-group_siz$Size/sum(group_siz$Size)
group_siz$Width<-4*group_siz$Prop

x<-c(-max(abs(all_gs$Z), na.rm=T),0,max(abs(all_gs$Z), na.rm=T))
x<-(x-min(x))/(max(x)-min(x))
  
heatmap<-ggplot(data = all_gs, aes(x = Panel, y = NAME)) +
	theme_bw()	+
	geom_tile(data=all_gs, aes(fill = Z), width=0.95, height=0.95) +
	geom_tile(data=all_gs[all_gs$P < 0.05,], aes(x = Panel, y = NAME), colour='black', fill=NA, size=0.3, width=0.95, height=0.95) +
	geom_tile(data=all_gs[all_gs$P.FDR < 0.05,], aes(x = Panel, y = NAME), colour='green2', fill=NA, size=0.3, width=0.95, height=0.95) +
	scale_fill_gradientn(colours=c("dodgerblue2","grey95","red"), na.value = NA,name = "Z-score", limits = c(-max(abs(all_gs$Z), na.rm=T), max(abs(all_gs$Z), na.rm=T)), values=x) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),plot.title = element_text(hjust = 0.5)) +
  labs(x='', y='') +
  facet_wrap(~ Method , nrow=1, scales = "free_x")

library(grid)
gt = ggplot_gtable(ggplot_build(heatmap))

for(i in 1:nrow(group_siz)){
  gt$widths[gt$layout$l[grep(paste0('panel-',i,'-1'), gt$layout$name)]] = group_siz$Width[i]*gt$widths[gt$layout$l[grep(paste0('panel-',i,'-1'), gt$layout$name)]]

}

png('/users/k1806347/oliverpainfel/Analyses/GCSC/drug_enrich_comp_magma_gcsc_twasgsea.png', units='px', res=300, width=2250, height=1200)
grid.draw(gt)
dev.off()

#######
# ATC associations
#######

###
# MAGMA
###

magma_gs_atc<-fread(paste0('/users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/results/ALS_only/magma/magma_drug_targetor_atc_res.csv'))
magma_gs_atc$P.FDR<-p.adjust(magma_gs_atc$P, method = 'fdr')
magma_gs_atc$log10_P<--log10(magma_gs_atc$P)
magma_gs_atc$FDR_Sig<-magma_gs_atc$P.FDR < 0.05
magma_gs_atc$Nom_Sig<-magma_gs_atc$P < 0.05
magma_gs_atc$Name<-paste0(magma_gs_atc$ATC,': ',magma_gs_atc$Name)
magma_gs_atc$Method<-'MAGMA'
magma_gs_atc$Panel<-'MAGMA'

magma_gs_atc<-magma_gs_atc[,c("Name","log10_P","FDR_Sig","Nom_Sig","Method","Panel"), with=F]

###
# GCSC
###

gcsc_gs_atc<-fread(paste0('/users/k1806347/oliverpainfel/Analyses/GCSC/GCSC_Brain_Blood_ALS_only_atc_res.csv'))
gcsc_gs_atc$log10_P<--log10(gcsc_gs_atc$P)
gcsc_gs_atc$FDR_Sig<-gcsc_gs_atc$P.FDR < 0.05
gcsc_gs_atc$Nom_Sig<-gcsc_gs_atc$P < 0.05
gcsc_gs_atc$Name<-paste0(gcsc_gs_atc$ATC,': ',gcsc_gs_atc$Name)
gcsc_gs_atc$Method<-'GCSC'
gcsc_gs_atc$Panel<-'GCSC'

gcsc_gs_atc<-gcsc_gs_atc[,c("Name","log10_P","FDR_Sig","Nom_Sig","Method","Panel"), with=F]

###
# TWAS-GSEA
###

weights<-read.table(paste0('/users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/results/ALS_only/twas/list_of_weights.txt'))$V1
weights<-weights[!grepl('SPLIC',weights)]

gsea_drugtargetor_atc_all<-NULL
for(i in weights){
  gsea_drugtargetor_atc<-fread(paste0('/users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/results/ALS_only/twas/drugtargetor/twas_gsea_',i,'_res_atc_res.csv'))
  gsea_drugtargetor_atc$Panel<-i
  # Calc. one-sided p
  gsea_drugtargetor_atc$P.oneside<-gsea_drugtargetor_atc$P/2
  gsea_drugtargetor_atc$P.oneside[gsea_drugtargetor_atc$Estimate < 0] <- 1-gsea_drugtargetor_atc$P.oneside[gsea_drugtargetor_atc$Estimate < 0]
  gsea_drugtargetor_atc$P.FDR<-p.adjust(gsea_drugtargetor_atc$P, method = 'fdr')
  gsea_drugtargetor_atc$P.FDR.onside<-p.adjust(gsea_drugtargetor_atc$P.onside, method = 'fdr')
  gsea_drugtargetor_atc_all<-rbind(gsea_drugtargetor_atc_all, gsea_drugtargetor_atc)
}

gsea_drugtargetor_atc_all$P.FDR_all<-p.adjust(gsea_drugtargetor_atc_all$P, method = 'fdr')
gsea_drugtargetor_atc_all$P.FDR.onside_all<-p.adjust(gsea_drugtargetor_atc_all$P.oneside, method = 'fdr')

gsea_gs_atc<-gsea_drugtargetor_atc_all
gsea_gs_atc$log10_P<-sign(gsea_gs_atc$Estimate)*-log10(gsea_gs_atc$P)
gsea_gs_atc$FDR_Sig<-gsea_gs_atc$P.FDR_all < 0.05
gsea_gs_atc$Nom_Sig<-gsea_gs_atc$P < 0.05
gsea_gs_atc$Name<-paste0(gsea_gs_atc$ATC,': ',gsea_gs_atc$Name)

gsea_gs_atc$Method<-'TWAS-GSEA'

gsea_gs_atc<-gsea_gs_atc[,c("Name","log10_P","FDR_Sig","Nom_Sig","Method","Panel"), with=F]

# Update labels
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Anterior_cingulate_cortex_BA24"]<-"Brain - Anterior cingulate cortex BA24 (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "psychencode"]<-"Brain - DLPFC (PsychENCODE)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Cerebellar_Hemisphere"]<-"Brain - Cerebellar hemisphere (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "NTR.BLOOD.RNAARR"]<-"Blood (NTR)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Cortex"]<-"Brain - Cortex (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Caudate_basal_ganglia"]<-"Brain - Caudate basal ganglia (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "YFS.BLOOD.RNAARR"]<-"Blood (YFS)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Hippocampus"]<-"Brain - Hippocampus (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Substantia_nigra"]<-"Brain - Substantia nigra (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Nucleus_accumbens_basal_ganglia"]<-"Brain - Nucleus accumbens basal ganglia (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Putamen_basal_ganglia"]<-"Brain - Putamen basal ganglia (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Frontal_Cortex_BA9"]<-"Brain - Frontal cortex BA9 (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Whole_Blood"]<-"Blood (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "CMC.BRAIN.RNASEQ"]<-"Brain - DLPFC (CMC)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Cerebellum"]<-"Brain - Cerebellum (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Spinal_cord_cervical_c-1"]<-"Brain - Spinal cord vervical c-1 (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Hypothalamus"]<-"Brain - Hypothalamus (GTEx)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "kcl_brainbank_motor_cortex"]<-"Brain - Motor cortex (KCL Brainbank)"
gsea_gs_atc$Panel[gsea_gs_atc$Panel == "Brain_Amygdala"]<-"Brain - Amygdala (GTEx)"

###
# Combine results and plot
###
all_gs_atc<-do.call(rbind, list(magma_gs_atc, gcsc_gs_atc, gsea_gs_atc))
all_gs_atc<-all_gs_atc[all_gs_atc$Name %in% all_gs_atc$Name[all_gs_atc$FDR_Sig],]

all_gs_atc$Method<-factor(all_gs_atc$Method, levels=c('MAGMA','GCSC','TWAS-GSEA'))

all_gs_atc$Name[grepl('C08C: selective calcium', all_gs_atc$Name)]<-'C08C: selective calcium channel blockers\nwith mainly vascular effects'
  
library(ggplot2)
library(cowplot)

group_siz<-NULL
for(i in c('MAGMA','GCSC','TWAS-GSEA')){
  group_siz<-rbind(group_siz, data.frame(Group=i,
                           Size=length(unique(all_gs_atc$Panel[all_gs_atc$Method==i]))))
}

# Set minimum size to 3 to allow space for labels
group_siz$Size[group_siz$Size < 2]<-2
group_siz$Prop<-group_siz$Size/sum(group_siz$Size)
group_siz$Width<-4*group_siz$Prop

x<-c(min(all_gs_atc$log10_P, na.rm=T),0,max(all_gs_atc$log10_P, na.rm=T))
x<-(x-min(x))/(max(x)-min(x))

heatmap<-ggplot(data = all_gs_atc, aes(x = Panel, y = Name)) +
	theme_bw()	+
	geom_tile(data=all_gs_atc, aes(fill = log10_P), width=0.95, height=0.95) +
	geom_tile(data=all_gs_atc[all_gs_atc$Nom_Sig == T,], aes(x = Panel, y = Name), colour='black', fill=NA, size=0.3, width=0.95, height=0.95) +
	geom_tile(data=all_gs_atc[all_gs_atc$FDR_Sig == T,], aes(x = Panel, y = Name), colour='green2', fill=NA, size=0.3, width=0.95, height=0.95) +
	scale_fill_gradientn(colours=c("blue","grey95","red"),name = "-log10(P)", limits = c(min(all_gs_atc$log10_P, na.rm=T), max(all_gs_atc$log10_P, na.rm=T)), values=x) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),plot.title = element_text(hjust = 0.5)) +
  labs(x='', y='') +
  facet_wrap(~ Method , nrow=1, scales = "free_x")

library(grid)
gt = ggplot_gtable(ggplot_build(heatmap))

for(i in 1:nrow(group_siz)){
  gt$widths[gt$layout$l[grep(paste0('panel-',i,'-1'), gt$layout$name)]] = group_siz$Width[i]*gt$widths[gt$layout$l[grep(paste0('panel-',i,'-1'), gt$layout$name)]]

}

png('/users/k1806347/oliverpainfel/Analyses/GCSC/atc_enrich_comp_magma_gcsc_twasgsea.png', units='px', res=300, width=3000, height=1750)
grid.draw(gt)
dev.off()

#######
# ATC level 4 associations
#######

###
# MAGMA
###

magma_gs_atc_level4<-fread(paste0('/users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/results/ALS_only/magma/magma_drug_targetor_atc_res_level4.csv'))
magma_gs_atc_level4$P.FDR<-p.adjust(magma_gs_atc_level4$P, method = 'fdr')
magma_gs_atc_level4$log10_P<--log10(magma_gs_atc_level4$P)
magma_gs_atc_level4$FDR_Sig<-magma_gs_atc_level4$P.FDR < 0.05
magma_gs_atc_level4$Nom_Sig<-magma_gs_atc_level4$P < 0.05
magma_gs_atc_level4$Name<-paste0(magma_gs_atc_level4$ATC,': ',magma_gs_atc_level4$Name)
magma_gs_atc_level4$Method<-'MAGMA'
magma_gs_atc_level4$Panel<-'MAGMA'

magma_gs_atc_level4<-magma_gs_atc_level4[,c("Name","log10_P","FDR_Sig","Nom_Sig","Method","Panel"), with=F]

###
# GCSC
###

gcsc_gs_atc_level4<-fread(paste0('/users/k1806347/oliverpainfel/Analyses/GCSC/GCSC_Brain_Blood_ALS_only_atc_res_level4.csv'))
gcsc_gs_atc_level4$log10_P<--log10(gcsc_gs_atc_level4$P)
gcsc_gs_atc_level4$FDR_Sig<-gcsc_gs_atc_level4$P.FDR < 0.05
gcsc_gs_atc_level4$Nom_Sig<-gcsc_gs_atc_level4$P < 0.05
gcsc_gs_atc_level4$Name<-paste0(gcsc_gs_atc_level4$ATC,': ',gcsc_gs_atc_level4$Name)
gcsc_gs_atc_level4$Method<-'GCSC'
gcsc_gs_atc_level4$Panel<-'GCSC'

gcsc_gs_atc_level4<-gcsc_gs_atc_level4[,c("Name","log10_P","FDR_Sig","Nom_Sig","Method","Panel"), with=F]

###
# TWAS-GSEA
###

weights<-read.table(paste0('/users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/results/ALS_only/twas/list_of_weights.txt'))$V1
weights<-weights[!grepl('SPLIC',weights)]

gsea_drugtargetor_atc_all_level4<-NULL
for(i in weights){
  gsea_drugtargetor_atc<-fread(paste0('/users/k1806347/oliverpainfel/Software/MyGit/GenoFunc/GenoFuncPipe/results/ALS_only/twas/drugtargetor/twas_gsea_',i,'_res_atc_res_level4.csv'))
  gsea_drugtargetor_atc$Panel<-i
  # Calc. one-sided p
  gsea_drugtargetor_atc$P.oneside<-gsea_drugtargetor_atc$P/2
  gsea_drugtargetor_atc$P.oneside[gsea_drugtargetor_atc$Estimate < 0] <- 1-gsea_drugtargetor_atc$P.oneside[gsea_drugtargetor_atc$Estimate < 0]
  gsea_drugtargetor_atc$P.FDR<-p.adjust(gsea_drugtargetor_atc$P, method = 'fdr')
  gsea_drugtargetor_atc$P.FDR.onside<-p.adjust(gsea_drugtargetor_atc$P.onside, method = 'fdr')
  gsea_drugtargetor_atc_all_level4<-rbind(gsea_drugtargetor_atc_all_level4, gsea_drugtargetor_atc)
}

gsea_drugtargetor_atc_all_level4$P.FDR_all<-p.adjust(gsea_drugtargetor_atc_all_level4$P, method = 'fdr')
gsea_drugtargetor_atc_all_level4$P.FDR.onside_all<-p.adjust(gsea_drugtargetor_atc_all_level4$P.oneside, method = 'fdr')

gsea_gs_atc_level4<-gsea_drugtargetor_atc_all_level4
gsea_gs_atc_level4$log10_P<-sign(gsea_gs_atc_level4$Estimate)*-log10(gsea_gs_atc_level4$P)
gsea_gs_atc_level4$FDR_Sig<-gsea_gs_atc_level4$P.FDR_all < 0.05
gsea_gs_atc_level4$Nom_Sig<-gsea_gs_atc_level4$P < 0.05
gsea_gs_atc_level4$Name<-paste0(gsea_gs_atc_level4$ATC,': ',gsea_gs_atc_level4$Name)

gsea_gs_atc_level4$Method<-'TWAS-GSEA'

gsea_gs_atc_level4<-gsea_gs_atc_level4[,c("Name","log10_P","FDR_Sig","Nom_Sig","Method","Panel"), with=F]

# Update labels
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Anterior_cingulate_cortex_BA24"]<-"Brain - Anterior cingulate cortex BA24 (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "psychencode"]<-"Brain - DLPFC (PsychENCODE)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Cerebellar_Hemisphere"]<-"Brain - Cerebellar hemisphere (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "NTR.BLOOD.RNAARR"]<-"Blood (NTR)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Cortex"]<-"Brain - Cortex (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Caudate_basal_ganglia"]<-"Brain - Caudate basal ganglia (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "YFS.BLOOD.RNAARR"]<-"Blood (YFS)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Hippocampus"]<-"Brain - Hippocampus (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Substantia_nigra"]<-"Brain - Substantia nigra (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Nucleus_accumbens_basal_ganglia"]<-"Brain - Nucleus accumbens basal ganglia (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Putamen_basal_ganglia"]<-"Brain - Putamen basal ganglia (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Frontal_Cortex_BA9"]<-"Brain - Frontal cortex BA9 (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Whole_Blood"]<-"Blood (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "CMC.BRAIN.RNASEQ"]<-"Brain - DLPFC (CMC)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Cerebellum"]<-"Brain - Cerebellum (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Spinal_cord_cervical_c-1"]<-"Brain - Spinal cord vervical c-1 (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Hypothalamus"]<-"Brain - Hypothalamus (GTEx)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "kcl_brainbank_motor_cortex"]<-"Brain - Motor cortex (KCL Brainbank)"
gsea_gs_atc_level4$Panel[gsea_gs_atc_level4$Panel == "Brain_Amygdala"]<-"Brain - Amygdala (GTEx)"

###
# Combine results and plot
###
all_gs_atc_level4<-do.call(rbind, list(magma_gs_atc_level4, gcsc_gs_atc_level4, gsea_gs_atc_level4))

# Retain only ATC codes 4 code within enriched ATC 3 codes.
all_gs_atc_level4<-all_gs_atc_level4[grepl(gsub(':.*','',all_gs_atc$Name[all_gs_atc$FDR_Sig == T]),all_gs_atc_level4$Name),]

all_gs_atc_level4$Method<-factor(all_gs_atc_level4$Method, levels=c('MAGMA','GCSC','TWAS-GSEA'))

library(ggplot2)
library(cowplot)

group_siz<-NULL
for(i in c('MAGMA','GCSC','TWAS-GSEA')){
  group_siz<-rbind(group_siz, data.frame(Group=i,
                           Size=length(unique(all_gs_atc_level4$Panel[all_gs_atc_level4$Method==i]))))
}

# Set minimum size to 3 to allow space for labels
group_siz$Size[group_siz$Size < 2]<-2
group_siz$Prop<-group_siz$Size/sum(group_siz$Size)
group_siz$Width<-4*group_siz$Prop

x<-c(min(all_gs_atc_level4$log10_P, na.rm=T),0,max(all_gs_atc_level4$log10_P, na.rm=T))
x<-(x-min(x))/(max(x)-min(x))

heatmap<-ggplot(data = all_gs_atc_level4, aes(x = Panel, y = Name)) +
	theme_bw()	+
	geom_tile(data=all_gs_atc_level4, aes(fill = log10_P), width=0.95, height=0.95) +
	geom_tile(data=all_gs_atc_level4[all_gs_atc_level4$Nom_Sig == T,], aes(x = Panel, y = Name), colour='black', fill=NA, size=0.3, width=0.95, height=0.95) +
	geom_tile(data=all_gs_atc_level4[all_gs_atc_level4$FDR_Sig == T,], aes(x = Panel, y = Name), colour='green2', fill=NA, size=0.3, width=0.95, height=0.95) +
	scale_fill_gradientn(colours=c("blue","grey95","red"),name = "-log10(P)", limits = c(min(all_gs_atc_level4$log10_P, na.rm=T), max(all_gs_atc_level4$log10_P, na.rm=T)), values=x) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),plot.title = element_text(hjust = 0.5)) +
  labs(x='', y='') +
  facet_wrap(~ Method , nrow=1, scales = "free_x")

library(grid)
gt = ggplot_gtable(ggplot_build(heatmap))

for(i in 1:nrow(group_siz)){
  gt$widths[gt$layout$l[grep(paste0('panel-',i,'-1'), gt$layout$name)]] = group_siz$Width[i]*gt$widths[gt$layout$l[grep(paste0('panel-',i,'-1'), gt$layout$name)]]

}

png('/users/k1806347/oliverpainfel/Analyses/GCSC/atc_level4_enrich_comp_magma_gcsc_twasgsea.png', units='px', res=300, width=3000, height=5000)
grid.draw(gt)
dev.off()
```

***

## Create plot comparing ATC level 4 enrichments for 