---
title: "GenoFuncPipe Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
```

* Download and format GWAS sumstats for brain related disorders
* Run pipeline for all GWAS
* Collate and compare results

***

# Download and format GWAS sumstats

***

### Schizophrenia (SCZ)

PMID: 35396580
EUR only

```{r, eval=F, echo=T}
dir.create('/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/SCZ', recursive = T)
system('wget --no-check-certificate -O /users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/SCZ/SCZ.original.gz https://figshare.com/ndownloader/files/34517828')

ss<-fread('/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/SCZ/SCZ.original.gz')

# Update column names. Use NEFF as N
# I am not sure how PGS is computing NEFF but appears wrong
names(ss)<-c('CHR','SNP','ORIGBP','A1','A2','FCAS','FCON','INFO','BETA','SE','P','NCAS','NCON','NEFF')
ss$N<-4/((1/ss$NCAS)+(1/ss$NCON))

# Calculate overall FREQ
ss$FREQ<-((ss$FCAS*ss$NCAS) + (ss$FCON*ss$NCON))/(ss$NCAS + ss$NCON)

# Keep only relevent columns
ss<-ss[,c('SNP','A1','A2','INFO','BETA','SE','P','N','FREQ'), with=F]

fwrite(ss, '/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/SCZ/SCZ.reformat.gz', sep=' ', na='NA', quote=F)

```

***

### Bipolar Disorder (BIP)

PMID: 34002096

```{r, eval=F, echo=T}
dir.create('/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/BIP', recursive = T)
system('wget --no-check-certificate -O /users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/BIP/BIP.original.gz https://figshare.com/ndownloader/files/26603681')

ss<-fread('/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/BIP/BIP.original.gz')

# Update column names. Use NEFF as N
# I am not sure how PGS is computing NEFF but appears wrong
names(ss)<-c('CHR','ORIGBP','SNP','A1','A2','BETA','SE','P','NGT','FCAS','FCON','INFO','NEFFDIV2','NCAS','NCON','DIR')
ss$N<-4/((1/ss$NCAS)+(1/ss$NCON))

# Calculate overall FREQ
ss$FREQ<-((ss$FCAS*ss$NCAS) + (ss$FCON*ss$NCON))/(ss$NCAS + ss$NCON)

# Keep only relevent columns
ss<-ss[,c('SNP','A1','A2','INFO','BETA','SE','P','N','FREQ'), with=F]

fwrite(ss, '/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/BIP/BIP.reformat.gz', sep=' ', na='NA', quote=F)

```

***

### Major Depressive Disorder (MDD)

I have access to the latest sumstat (incl. 23andMe). Not yet published.

```{r, eval=F, echo=T}

dir.create('/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/MDD', recursive = T)

# Read in the sumstats
ss<-fread('/users/k1806347/brc_scratch/Software/mdd-meta_2/mdd-meta/results/distribution/daner_pgc_mdd_full_eur_hg19_v3.49.24.11.gz')

# Update column names
names(ss)<-c('CHR','SNP','ORIGBP','A1','A2','FCAS','FCON','INFO','OR','SE','P','NGT','DIR','HETISQT','HETDF','HETP','NCAS','NCON','NEFFDIV2')
ss$N<-4/((1/ss$NCAS)+(1/ss$NCON))

# Calculate overall FREQ
ss$FREQ<-((ss$FCAS*ss$NCAS) + (ss$FCON*ss$NCON))/(ss$NCAS + ss$NCON)

# Keep only relevent columns
ss<-ss[,c('SNP','A1','A2','INFO','OR','SE','P','N','FREQ'), with=F]

fwrite(ss, '/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/MDD/MDD.reformat.gz', sep=' ', na='NA', quote=F)

```

***

### ADHD

PMID: 30478444
Already available on Rosalind: /scratch/groups/ukbiobank/sumstats/cleaned/ADHD05.gz

***

### Autism Spectrum Disorders (ASD)

PMID: 30804558
Already available on Rosalind: /scratch/groups/ukbiobank/sumstats/cleaned/AUTI07.gz

***

### Amyotrophic Lateral Sclerosis (ALS)

PMID: 34873335
European only.
Already available on Rosalind: /users/k1806347/brc_scratch/Data/GWAS_sumstats/ALS/GWAS/ALS_sumstats_EUR_only_reformat.txt.gz

***

### Alzheimer's Disease (ALZ)

PMID: 34493870
Excludes 23andMe.

```{r, eval=F, echo=T}
dir.create('/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/ALZ', recursive = T)
system('wget --no-check-certificate -O /users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/ALZ/ALZ.original.gz https://ctg.cncr.nl/documents/p1651/PGCALZ2sumstatsExcluding23andMe.txt.gz')

ss<-fread('/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/ALZ/ALZ.original.gz')

# Update column names.
names(ss)<-c('CHR','ORIGBP','A1','A2','BETA','P','N')

fwrite(ss, '/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/ALZ/ALZ.reformat.gz', sep=' ', na='NA', quote=F)

```


***

### Parkinson's Disease (PRK)

Publicly available version
PMID: 31701892

```{r, eval=F, echo=T}
dir.create('/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/PRK', recursive = T)
system('wget --load-cookies /tmp/cookies.txt "https://docs.google.com/uc?export=download&confirm=$(wget --quiet --save-cookies /tmp/cookies.txt --keep-session-cookies --no-check-certificate 'https://docs.google.com/uc?export=download&id=1FZ9UL99LAqyWnyNBxxlx6qOUlfAnublN' -O- | sed -rn 's/.*confirm=([0-9A-Za-z_]+).*/\1\n/p')&id=1FZ9UL99LAqyWnyNBxxlx6qOUlfAnublN" -O /users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/PRK/PRK.original.zip && rm -rf /tmp/cookies.txt')
setwd('/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/PRK/')
system('unzip /users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/PRK/PRK.original.zip')
system('rm /users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/PRK/PRK.original.zip')

ss<-fread('/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/PRK/nallsEtAl2019_excluding23andMe_allVariants.tab')

ss$CHR<-gsub('chr','',gsub(':.*','',ss$SNP))
ss$BP<-gsub('.*:','',ss$SNP)

# Update column names.
names(ss)<-c('SNP','A1','A2','FREQ','BETA','SE','P','NCAS','NCON','CHR','ORIGBP')

ss$N<-4/((1/ss$NCAS)+(1/ss$NCON))

# Keep only relevent columns
ss<-ss[,c('CHR','ORIGBP','A1','A2','BETA','SE','P','N','FREQ'), with=F]

fwrite(ss, '/users/k1806347/brc_scratch/Data/GWAS_sumstats/GenoFuncPipe_paper/PRK/PRK.reformat.gz', sep=' ', na='NA', quote=F)

```

***

### Migraine (MIG)

PMID: 35115687
Excl. 23andMe
Requested from Dale Nyholt.

***

## Run pipeline for all GWAS

I have made a list of GWAS called gwas_list_brain.txt
I have made a config file specifying that GWAS list called config_brain.yaml
I have adapted the Snakefile to specify that config file

```{bash, eval=F, echo=T}

snakemake -n --use-conda --profile slurm run_create_report

```

***

