---
title: "Validating the ATC drug-class enrichment estimator: correlation diagnostics, calibration, and a positive-control benchmark"
author: "GenoDisc pipeline"
date: "2026-09-28"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressMessages(library(data.table))
kab <- function(x, ...) knitr::kable(x, ...)
```

> **Scope.** This is an *analysis/validation* report. It does not itself change the
> pipeline. All work was done in a scratch directory against existing job outputs;
> the scripts and job directories are listed in the [Reproducibility](#reproducibility)
> appendix. Data throughout use **unconditioned** per-drug statistics and an
> **unconditioned** drug–drug correlation matrix (statistics and the correlation
> matrix always matched in conditioning).

# Summary {-}

The drug-level GLS ATC test (DRUGSETS-style, eigenvalue-floor pseudo-inverse) gave
pathological rankings on schizophrenia: small, internally-mixed classes floated to
the top (MAGMA `C03C` high-ceiling diuretics; TWAS-GSEA `S01E` antiglaucoma) while a
large, coherent, strongly-associated class (`N05A` antipsychotics) was buried. This
report diagnoses why and compares estimators.

Findings:

1. **The GLS's anti-conservatism was a residual-degrees-of-freedom bug**, not a
   modelling subtlety. A rank-\(m\) pseudo-inverse projects the residual into an
   \(m\)-dimensional subspace, so the residual variance must be divided by \(m-K\),
   not \(N-K\). Fixed, the correct-model type-I error is nominal for every estimator.
2. **The df fix corrects p-values but not the ranking.** The ordering pathology is
   intrinsic to the correlation-whitening pseudo-inverse: it **inflates small mixed
   classes via out-of-class leakage and dilutes large coherent ones**. We demonstrate
   this explicitly for `C03C` (111% of its coefficient comes from *non-member* drugs).
3. **Recommendation: use CAMERA-style variance-inflated OLS (VIF-OLS)** as the ATC
   enrichment estimator for both engines, keeping the legacy Wilcoxon as a secondary
   cross-check. The justification is the **estimand** and the **demonstrated GLS
   leakage**, *not* a sign-agreement scorecard (see caveats).
4. **The strong control validates the engine choice, not the drug-targetor TWAS.**
   Lipid-lowering (`C10A`) on coronary artery disease is recovered on **MAGMA** at
   rank 2, FDR 0.008 under VIF-OLS, but is **null on TWAS** for both estimators.

# 1. Motivation

The per-ATC drug-level test asks: *are the drugs in an ATC class more associated
(MAGMA) / more directionally concordant (TWAS-GSEA) with the trait than other drugs?*
Because drugs that share target genes have correlated per-drug statistics, a naive
test (the legacy competitive Wilcoxon) is anti-conservative. The DRUGSETS approach
(Bell et al. 2022, medRxiv 2022.09.06.22279660) addresses this with a GLS that
whitens by the drug–drug correlation \(\Sigma\), inverted keeping eigenvalues
\(\ge 0.1\).

On SCZ this produced rankings that did not survive scrutiny: `C03C` (6 loop
diuretics, internally mixed) ranked 2nd on MAGMA while `N05A` (48 antipsychotics)
was 23rd. We set out to determine whether \(\Sigma\) is mis-specified, whether the
estimator is miscalibrated, and which estimator recovers positive controls sensibly.

# 2. Data and scope

* **Engines / statistics.**
  MAGMA: per-drug competitive `BETA/SE` (non-directional magnitude), `size = NGENES`.
  TWAS-GSEA: per-drug signed `T` (directional), `size = N_Mem_Avail`.
* **Correlation matrix \(\Sigma\).** For each engine, the exact null correlation of
  the per-drug statistics (`cov2cor(Z_wh' Z_wh)` for TWAS; the analogous whitened
  membership correlation for MAGMA), aligned to the same drug set as the statistics.
* **Classes.** Every ATC L3 (4-char) class with \(\ge 5\) member drugs; multi-code
  drugs exploded into all their classes.
* **Calibration + benchmark GWAS.** SCZ (SCHI04) for the deep dive; a 6-trait
  benchmark — SCZ, bipolar (BIPO04), depression (DEPR02), ADHD (ADHD06), T2D
  (DIAB05), asthma (ASTH02) — on tissue-matched GTEx panels + whole blood; and CAD
  (COAD01) for the lipid strong control.

# 3. Estimators and the directional sign convention

Three estimators, all fitting `y ~ [1, class_indicator, size, log(size)]`:

* **VIF-OLS (CAMERA-style).** OLS coefficient on the class indicator with the
  standard error inflated for within-class correlation,
  \(\mathrm{se} = \mathrm{se}_{\mathrm{OLS}}\sqrt{1+(n_1-1)\bar\rho}\), where
  \(\bar\rho\) is the mean in-class entry of \(\Sigma\). It uses **only** the
  in-class-vs-rest contrast.
* **GLS, floor 0.1 (current default).** \(\Sigma^{-1}\) is a rank-\(m\)
  pseudo-inverse dropping eigenvalues \(<0.1\).
* **GLS, ridge.** \((\Sigma + \tau^2 I)^{-1}\), \(\tau^2\) by profile REML under the
  null. (Dropped after calibration; see §6.)

Tail conventions match production: **MAGMA one-sided** (enrichment), **TWAS
two-sided**.

## 3.1 The TWAS directional sign, traced end-to-end

To state what a negative coefficient means, we traced the sign through the code:

```{r sign-trace, eval=FALSE}
# (1) per-gene drug direction, format_drug_targetor_for_twas_gsea.R:28
#     -1 if the drug DECREASES the gene (DECREASED_EXPRESSION / NEGATIVE_RESPONSE /
#         OPPOSITE_RESPONSE); +1 if it INCREASES it (INCREASED/POSITIVE_RESPONSE); else 0
# (2) outcome, TWAS-GSEA-fast.R:176-177  (invoked with --directional T)
#     TWAS$ZSCORE <- TWAS$TWAS.Z          # SIGNED gene TWAS z (not the |z| default)
# (3) statistic, TWAS-GSEA-fast.R:427-431
#     both the +/-1 prop and signed z are whitened by the gene-gene correlation and
#     residualised on [1, GeneLength, NSNP]; T = cov(whitened +/-1, whitened z) / sd
```

So the **per-drug \(T \propto \mathrm{cov}(\text{drug's } \pm1 \text{ direction},\
\text{signed gene TWAS } z)\)**, and the class coefficient \(B\) is that contrast for
in-class vs out-of-class drugs. Therefore:

* **\(B>0\) (positive coefficient):** in-class drugs move gene expression
  **concordantly** with the disease TWAS signature (code label: "Matches disease").
* **\(B<0\) (negative coefficient):** in-class drugs move expression **discordantly**
  with (opposite to) the disease signature (code label: "Opposes disease";
  `Reversal_Z = -T`).

> **Terminology.** We deliberately report **"positive/negative coefficient"
> (concordant/discordant)** and *not* "therapeutic/reversal". Reading a discordant
> signature as therapeutic benefit is an additional, unvalidated assumption (correct
> orientation of the disease \(z\), completeness of the drug-action annotations, and
> that signature reversal is beneficial). MAGMA is non-directional: \(B>0\) =
> "Enriched", \(B<0\) = "Depleted".

# 4. The correlation matrix \(\Sigma\)

```{r R-diag}
Rdiag <- data.table(
  Engine        = c("MAGMA (SCZ)", "TWAS-GSEA BA9 (SCZ)"),
  n_drugs       = c(1446L, 678L),
  `m = #lambda>=0.1` = c(1033L, 438L),
  `min(lambda)` = c("-2.5e-15", "-1.2e-14"),
  `drugs w/ an exact duplicate target set` = c("112 (34 clusters)", "64 (19 clusters)"),
  `#|lambda|<1e-8` = c(147L, 151L)
)
kab(Rdiag, caption = "Sigma is singular. Exact-duplicate target sets (perfectly correlated drugs) account for roughly half the rank deficiency; the rest is near-collinearity.")
```

An eigenvalue **whitening diagnostic** — the variance of
\(u_k = v_k^\top r / \sqrt{\lambda_k}\) across eigen-directions, which should be flat
at 1 if \(\Sigma\) is well specified — showed \(\mathrm{var}(u_k)\approx 1\) for
\(\lambda \gtrsim 0.05\) but **rising sharply in the small-\(\lambda\) tail** (up to
~25× on MAGMA, ~2–3× on TWAS). So \(\Sigma\) understates the true variance in its
small-eigenvalue directions, exactly where a naive \(1/\lambda\) pseudo-inverse
over-weights. This motivated both the floor and the calibration below.

# 5. The residual-degrees-of-freedom bug

The production `gls_group` estimates the residual variance as
\(\hat\sigma^2 = r^\top \Sigma^{+} r / (N-K)\) and tests with \(N-K\) degrees of
freedom. But \(\Sigma^{+}\) is a **rank-\(m\)** pseudo-inverse (\(m=\#\{\lambda\ge0.1\}\)),
so it projects the residual into an \(m\)-dimensional subspace:
\[
r^\top \Sigma^{+} r \;\sim\; \sigma^2 \,\chi^2_{\,m-K}.
\]
Dividing by \(N-K\) (with \(N\gg m\)) under-estimates \(\sigma^2\), inflating the
t-statistic. The fix is to use \(m-K\) in both the variance and the reference
distribution:

```{r df-fix, eval=FALSE}
sigma2 <- as.numeric(t(resid) %*% Sinv %*% resid) / (m - K)   # was (N - K)
tval   <- B[2] / sqrt(sigma2 * W[2, 2]); df <- m - K          # was N - K
```

# 6. Null calibration

Type-I error was estimated by MVN simulation (1000 reps; ~126k tests MAGMA, ~68k
TWAS). Two requirements were enforced: (i) each estimator is first checked under a
simulation drawn from **its own** assumed covariance (the "correct-model gate"), and
(ii) the estimator and the simulation use the **same** matrix (the ridge is only
well-defined once \(\tau^2\) is floored so \(\Sigma+\tau^2 I\) is positive definite).

```{r calib}
cal <- data.table(
  Scenario = c("Correct-model gate","Correct-model gate","Correct-model gate","Correct-model gate",
               "Empirical tail","Empirical tail","Empirical tail","Empirical tail"),
  Engine   = c("MAGMA","MAGMA","TWAS BA9","TWAS BA9","MAGMA","MAGMA","TWAS BA9","TWAS BA9"),
  Metric   = rep(c("VIF-OLS","GLS floor-0.1 (df-fixed)"), 4),
  `type-I @0.05` = c(0.038, 0.050, 0.042, 0.050, 0.038, 0.053, 0.042, 0.048)
)
kab(cal, caption = "After the df fix, the correct-model type-I is nominal for VIF-OLS and the GLS. Ridge also passed its correct-model gate (0.051 / 0.050) but was badly conservative under the realistic empirical-tail on the singular TWAS matrix (0.0038) and was dropped.")
```

* **VIF-OLS:** robust, mildly conservative everywhere.
* **GLS floor-0.1 (df-fixed):** now well-calibrated even under the realistic
  empirical tail — it discards the mis-specified small-\(\lambda\) directions.
* **Ridge:** fragile where \(\Sigma\) is singular and REML wants \(\tau^2\to0\)
  (TWAS); dropped.

**Key consequence:** the df fix rescales p-values by a per-engine constant, so it
does **not** change the class *ranking*. The ordering pathology (§8) is a property
of the point estimate, and is only settled by the benchmark.

# 7. Positive-control benchmark

Pre-specified controls (expected: MAGMA Enriched, TWAS concordant/discordant per the
drug mechanism), scored on P, FDR (BH within trait×engine), sign, and rank.

```{r bench-controls}
ctrl <- data.table(
  Trait = c("SCZ","SCZ","BIP","BIP","DEP","ADHD","ADHD","T2D","T2D","ASTH","ASTH"),
  Class = c("N05A","N05A","N03A","N05A","N06A","N06B","N06B","A10B","A10B","R03B","R03B"),
  Engine= c("MAGMA","TWAS BA9","MAGMA","TWAS cortex","MAGMA","MAGMA","TWAS cortex","MAGMA","TWAS pancreas","MAGMA","TWAS lung"),
  `VIF rank` = c(5,10,2,66,48,8,39,1,16,24,8),
  `GLS rank` = c(23,24,54,72,101,1,22,3,6,107,37)
)
kab(ctrl, caption = "Representative pre-specified controls (rank among all L3 classes; lower is better). Head-to-head across all controls is genuinely MIXED (VIF better on 11, GLS on 13).")
```

```{r bench-nsig}
nsig <- data.table(
  Estimator = c("VIF-OLS","GLS floor-0.1 (df-fixed)"),
  `FDR<0.05 classes across all 18 runs` = c("4  (SCZ-MAGMA: C08C, N03A;  DEP-cortex: C05A;  ASTH-blood: R03D)", "0")
)
kab(nsig, caption = "FDR-significant classes. GLS found none in any run.")
```

> **Reporting the metrics honestly.** The head-to-head control rank is **mixed**
> (11 VIF / 13 GLS), and GLS is better on whole-blood panels (the wrong tissue). We
> **do not** headline the directional sign-agreement comparison: with only 8
> tissue-matched TWAS controls, all non-significant, it is underpowered. The case for
> switching rests on the estimand and the mechanism below, not on this scorecard.

# 8. Why the GLS misranks: out-of-class leakage

Decomposing the GLS coefficient as \(B = \sum_i w_i y_i\) (effective weight × per-drug
statistic) exposes the mechanism. For the flagged `C03C`:

```{r c03c}
c03c <- data.table(
  drug = c("BUMETANIDE","TORASEMIDE","CHLORTALIDONE","QUINETHAZONE","METHYCLOTHIAZIDE","PIRETANIDE","BENDROFLUMETHIAZIDE"),
  in_class = c("IN","IN","out","out","out","IN","out"),
  y = c(1.91,-1.96,-2.05,-1.55,-2.11,-0.55,-2.07),
  contribution = c(0.434,-0.374,0.234,0.195,0.134,-0.116,0.110)
)
kab(c03c, caption = "SCZ MAGMA C03C (GLS rank 2, VIF rank 55). Top contributors to the GLS coefficient.")
```

The class coefficient is \(B_{\mathrm{GLS}} = 0.766\), of which the **in-class share
is \(-0.083\)** and the **out-of-class share is \(+0.849\)** — i.e. **111% of the
"enrichment" comes from non-member drugs**. The six loop diuretics roughly cancel;
correlated out-of-class thiazides (`C03A`/`C03B`), which are *depleted* (\(y<0\)),
receive negative pseudo-inverse weights and contribute positively. `C03C` looks
enriched only because its correlated non-members are depleted.

By contrast the true control **`N05A`** is *diluted*: \(B_{\mathrm{GLS}}=0.13\)
(rank 23) vs \(B_{\mathrm{VIF}}=0.90\) (rank 5) — 48 coherent drugs each receive
tiny weights while a 6-drug class gets weights up to 0.23. This small-class-inflation
/ large-class-dilution is **systematic**: 79 class×run cases where GLS ranks a class
top-10 while VIF ranks it >30, overwhelmingly small classes. VIF-OLS, using only the
in-class-vs-rest contrast, is immune.

# 9. VIF's FDR-significant classes, with leave-one-gene-out

For each of VIF's four FDR-significant classes we identified the top contributing
drugs and genes, and asked whether a single gene dominates (an attribution-level
leave-one-gene-out on the gene-contribution decomposition: signed membership × gene
association — the same quantity shown in the Shiny evidence panel).

```{r vif-fdr}
vf <- data.table(
  Class = c("C08C","N03A","C05A","R03D"),
  Engine = c("SCZ / MAGMA","SCZ / MAGMA","DEP / TWAS cortex","ASTH / TWAS blood"),
  N = c(15,37,13,9),
  B = c(1.68,1.16,-1.10,2.11),
  FDR = c(0.027,0.027,0.025,0.012),
  `top genes` = c("CACNA1C, CACNB2, CACNA1S", "CACNA1I, SCN9A, SCN1A (over 7718 genes)",
                  "FLOT1, DFNA5, AK2 (diffuse)", "ADORA1, MDH1, IL6R"),
  `single-gene dominance` = c("no (top 9%, top-3 44%)","no (top 1%)","no (top 3%)","no (top 3%, top-3 22%)")
)
kab(vf, caption = "No single gene dominates any class. C08C is a coherent calcium-channel target/risk-gene overlap (CACN* family); R03D reflects the xanthine adenosine-receptor targets; N03A and C05A are diffuse. C05A (a non-psychiatric class) is most consistent with a chance FDR hit.")
```

# 10. Strong control: lipid-lowering on CAD

`C10A` (statins, fibrates, ezetimibe) on coronary artery disease is the canonical
strong control. MAGMA drug-targetor was regenerated for the CAD job to test it on
both engines.

```{r c10a}
c10a <- data.table(
  Engine = c("MAGMA","MAGMA","TWAS Artery_Aorta","TWAS Whole_Blood"),
  Estimator = c("VIF-OLS","GLS floor-0.1","VIF-OLS","VIF-OLS"),
  B = c(1.44, 0.68, -0.47, 0.02),
  P = c("1.3e-4","1.4e-3","0.22","0.96"),
  FDR = c("0.008","0.085","n.s.","n.s."),
  rank = c("2 / 126","2 / 126","25 / 90","94 / 96")
)
kab(c10a, caption = "C10A on CAD. Recovered on MAGMA (VIF rank 2, FDR 0.008; drivers HMGCR/LDLR/APOB/LPL, no single-gene dominance). GLS dilutes it below FDR. Null on TWAS for both estimators.")
```

The full CAD/MAGMA VIF top-3 is cardiovascular-coherent: `C01D` (nitrates, rank 1),
`C10A` (rank 2), `C02K` (antihypertensives, rank 3). This shows the drug-targetor
signal **can** appear cleanly — on MAGMA. The TWAS drug-targetor is the weak engine,
independent of estimator.

# 11. Two alignment bugs to fix alongside the switch

1. **`drugcorr.rds` vs `competitive.clean.csv` drug-set mismatch.** For SCZ, both have
   707 rows but only 678 CIDs intersect (29 dropped each way). The correlation matrix
   and the statistics must be built on an identical, verified drug set.
2. **Primary-ATC-only keying.** The production TWAS GLS keys \(\Sigma\) by `res$ATC`
   (a single 7-char code, matched 1:1), which silently drops multi-ATC drugs and never
   explodes a drug into all its classes. Alignment should be by CID / full string with
   multi-code explosion (as done throughout this analysis).

# 12. Conclusion and recommendation

* **Switch the ATC drug-class enrichment estimator to VIF-OLS for both engines**,
  keeping the legacy Wilcoxon as a secondary cross-check. Justification: the estimand
  (in-class vs rest, correctly variance-inflated) and the demonstrated GLS
  out-of-class leakage — **not** the sign scorecard.
* **Apply the residual-df handling** if any pseudo-inverse GLS is retained anywhere
  (it corrects calibration, not ranking).
* **Fix the two alignment bugs** as part of the switch.
* **Frame results honestly.** Power is modest: apart from the CAD/MAGMA lipid control,
  most controls do not reach FDR significance under either estimator. Report enrichment
  as exploratory, describe direction as a **positive/negative coefficient**
  (concordant/discordant with the disease signature), and avoid a "reverses disease →
  therapeutic" claim until it is separately validated. **MAGMA is the stronger engine**
  for drug-targetor; the TWAS drug-targetor should remain labelled experimental.

# Reproducibility {#reproducibility}

Analysis scripts (run with system `/usr/bin/Rscript`; base R + `data.table`):

```{r repro, eval=FALSE}
# scratchpad/atc_gls_diag/
#   diagnose.R     - Step 1: eigenvalue whitening diagnostic
#   step23b.R      - residual-df fix, matched sim/estimator, correct-model gate,
#                    empirical-tail calibration  (-> step3b_type1.csv)
#   step4.R        - 6-trait x 2-engine positive-control benchmark
#                    (-> step4_controls.csv, step4_nsig.csv, step4_allclasses.csv)
#   item1.R        - VIF FDR-sig classes: contributing drugs/genes + leave-one-gene-out
#   item3.R        - C10A on CAD, TWAS panels
#   item3_magma.R  - C10A on CAD, MAGMA (-> coad01_magma_allclasses.csv)
#   preflight.R    - TWAS drug-count chain + duplicate-target-set counts
#
# Job directories (unconditioned stats + unconditioned Sigma):
#   jobs/claude_poscontrol_{SCHI04,BIPO04,DEPR02,ADHD06,DIAB05,ASTH02}_2026092*
#   jobs/claude_gls_coad01_gw_20260926_152549   (CAD; MAGMA drugtargetor regenerated)
#
# Production references (unchanged by this analysis):
#   scripts/format_{magma,twas_gsea}_drugtargetor_gls_results.R
#   scripts/format_drug_targetor_for_twas_gsea.R   (the +/-1 sign construction)
#   {resdir}/software/TWAS-GSEA/TWAS-GSEA-fast.R    (--directional; the per-drug T)
```
