---
title: "GenoFunc"
params:
  gwas: ""
output:
  html_document:
    toc: true
    theme: united
    toc_depth: 3
    toc_float: true
    fig_caption: yes

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../')

library(data.table)
library(knitr)
library(ggplot2)
library(cowplot)
library(kableExtra)

knit_hooks$set(optipng = hook_optipng)
knit_hooks$set(pngquant = hook_pngquant)

```

***

<style type="text/css">

h1 { /* Header 1 */
  font-size: 28px;
}
h2 { /* Header 2 */
  font-size: 22px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
</style>

```{css, echo=F}
pre code, pre, code {
  white-space: pre !important;
  overflow-x: auto !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
```

<font size="4"> 
ID: `r params$gwas`
</font>

***

# GWAS QC

<details><summary>Show sumstat_cleaner.R log file</summary>

```{r, eval=T, echo=F, comment=''}
sumstat_cleaner_log<-readLines(paste0('resources/data/gwas_sumstat/',params$gwas,'/',params$gwas,'.cleaned.log'))

sumstat_cleaner_log

```

</details>

<details><summary>Show FOCUS munge log file</summary>

```{r, eval=T, echo=F, comment=''}
focus_munge_log<-readLines(paste0('resources/data/gwas_sumstat/',params$gwas,'/',params$gwas,'.cleaned.munged.log'))

focus_munge_log

```

</details>

***

# Molecular associations

***

## TWAS

***

### PsychENCODE DLPFC

<details><summary>Show FDR significant and colocalised results</summary>

```{r, eval=T, echo=F}
twas_psychencode<-fread(paste0('results/',params$gwas,'/twas/psychencode/',params$gwas,'_twas_psychencode_GW'))

twas_psychencode$TWAS.P.FDR<-p.adjust(twas_psychencode$TWAS.P, method = 'fdr')
twas_psychencode<-twas_psychencode[,c('CHR','P0','P1','ID','TWAS.Z','TWAS.P','TWAS.P.FDR','COLOC.PP3','COLOC.PP4'), with=T]

twas_psychencode<-twas_psychencode[which(twas_psychencode$TWAS.P.FDR < 0.05 & twas_psychencode$COLOC.PP4 > 0.8),]

twas_psychencode<-twas_psychencode[order(twas_psychencode$CHR, twas_psychencode$P0),]

twas_psychencode$TWAS.Z<-round(twas_psychencode$TWAS.Z, 3)
twas_psychencode$TWAS.P<-format(twas_psychencode$TWAS.P, scientific = TRUE, digits = 3)
twas_psychencode$TWAS.P.FDR<-format(twas_psychencode$TWAS.P.FDR, scientific = TRUE, digits = 3)

twas_psychencode %>% 
  kable(booktabs = T) %>%
  kable_styling()

```

</details>

***

## MAGMA

<details><summary>Show FDR significant results</summary>

```{r, eval=T, echo=F}
magma<-fread(paste0('results/',params$gwas,'/magma/magma_gene_level.clean.csv'))

magma$P.FDR<-p.adjust(magma$P, method = 'fdr')

magma<-magma[which(magma$P.FDR < 0.05),]

magma<-magma[order(magma$CHR, magma$START),]

magma$P<-format(magma$P, scientific = TRUE, digits = 3)
magma$P.FDR<-format(magma$P.FDR, scientific = TRUE, digits = 3)

magma %>% 
  kable(booktabs = T) %>%
  kable_styling()

```

</details>

***

# Drug-repruposing

***

## TWAS-CMAP: So et al. Average Ranks Method

<details><summary>Show Top Drug Enrichment Results</summary>

```{r, eval=T, echo=F}
so_cmap<-fread(paste0('results/',params$gwas,'/twas/cmap/So_res_clean.csv'))
so_cmap<-so_cmap[order(so_cmap$p),]
  
if(sum(so_cmap$p.fdr < 0.05) >= 10){
  so_cmap<-so_cmap[so_cmap$p.fdr < 0.05,]
} else {
  so_cmap<-so_cmap[1:10,]
}

so_cmap$p<-format(so_cmap$p, scientific = TRUE, digits = 3)
so_cmap$p.fdr<-format(so_cmap$p.fdr, scientific = TRUE, digits = 3)

so_cmap<-so_cmap[,c("pert_mfc_id","cmap_name","pert_idose","pert_itime","cell_iname","avg.rank","p","p.fdr"), with=F]
names(so_cmap)<-c('ID','Name','Dose','Time','Cell','Rank','P','P.FDR')

so_cmap %>% 
  kable(booktabs = T) %>%
  kable_styling()

```

</details>

<details><summary>Show ATC Enrichment Results</summary>

```{r, eval=T, echo=F}
so_cmap_atc<-fread(paste0('results/',params$gwas,'/twas/cmap/So_res_atc_res.csv'))
so_cmap_atc<-so_cmap_atc[order(so_cmap_atc$P),]
so_cmap_atc$P.FDR<-p.adjust(so_cmap_atc$P, method = 'fdr')
  
if(sum(so_cmap_atc$P.FDR < 0.05) >= 10){
  so_cmap_atc<-so_cmap_atc[so_cmap_atc$P.FDR < 0.05,]
} else {
  so_cmap_atc<-so_cmap_atc[1:10,]
}

so_cmap_atc$P<-format(so_cmap_atc$P, scientific = TRUE, digits = 3)
so_cmap_atc$P.FDR<-format(so_cmap_atc$P.FDR, scientific = TRUE, digits = 3)

so_cmap_atc<-so_cmap_atc[,c("ATC","N","P","P.FDR","Name"), with=F]
names(so_cmap_atc)<-c("ATC","N Genes","P","P.FDR","Description")

so_cmap_atc %>% 
  kable(booktabs = T) %>%
  kable_styling()

```

</details>

***

## TWAS-CMAP: TWAS-GSEA Method

<details><summary>Show Top Drug Enrichment Results</summary>

```{r, eval=T, echo=F}
gsea_cmap<-fread(paste0('results/',params$gwas,'/twas/cmap/twas_gsea_cmap.competitive.clean.csv'))
gsea_cmap<-gsea_cmap[order(gsea_cmap$P),]
  
if(sum(gsea_cmap$P.CORR < 0.05) >= 10){
  gsea_cmap<-gsea_cmap[gsea_cmap$P.CORR < 0.05,]
} else {
  gsea_cmap<-gsea_cmap[1:10,]
}

gsea_cmap$P<-format(gsea_cmap$P, scientific = TRUE, digits = 3)
gsea_cmap$P.CORR<-format(gsea_cmap$P.CORR, scientific = TRUE, digits = 3)

gsea_cmap<-gsea_cmap[,c("pert_mfc_id","cmap_name","pert_idose","pert_itime","cell_iname","Estimate","SE","P","P.CORR") , with=F]
names(gsea_cmap)<-c('ID','Name','Dose','Time','Cell','Estimate','SE','P','P.FDR')

gsea_cmap %>% 
  kable(booktabs = T) %>%
  kable_styling()

```

</details>

<details><summary>Show ATC Enrichment Results</summary>

```{r, eval=T, echo=F}
gsea_cmap_atc<-fread(paste0('results/',params$gwas,'/twas/cmap/twas_gsea_res_atc_res.csv'))
gsea_cmap_atc<-gsea_cmap_atc[order(gsea_cmap_atc$P),]
gsea_cmap_atc$P.FDR<-p.adjust(gsea_cmap_atc$P, method = 'fdr')
  
if(sum(gsea_cmap_atc$P.FDR < 0.05) >= 10){
  gsea_cmap_atc<-gsea_cmap_atc[gsea_cmap_atc$P.FDR < 0.05,]
} else {
  gsea_cmap_atc<-gsea_cmap_atc[1:10,]
}

gsea_cmap_atc$P<-format(gsea_cmap_atc$P, scientific = TRUE, digits = 3)
gsea_cmap_atc$P.FDR<-format(gsea_cmap_atc$P.FDR, scientific = TRUE, digits = 3)

gsea_cmap_atc<-gsea_cmap_atc[,c("ATC","N","P","P.FDR","Name"), with=F]
names(gsea_cmap_atc)<-c("ATC","N Genes","P","P.FDR","Description")

gsea_cmap_atc %>% 
  kable(booktabs = T) %>%
  kable_styling()

```

</details>

***

## MAGMA-DrugTargetor

<details><summary>Show Top Drug Enrichment Results</summary>

```{r, eval=T, echo=F}
magma_gs<-fread(paste0('results/',params$gwas,'/magma/magma_drug_targetor.clean.csv'))
magma_gs<-magma_gs[order(magma_gs$P),]
magma_gs$P.FDR<-p.adjust(magma_gs$P, method='fdr')
  
if(sum(magma_gs$P.FDR < 0.05) >= 10){
  magma_gs<-magma_gs[magma_gs$P.FDR < 0.05,]
} else {
  magma_gs<-magma_gs[1:10,]
}

magma_gs$P<-format(magma_gs$P, scientific = TRUE, digits = 3)
magma_gs$P.FDR<-format(magma_gs$P.FDR, scientific = TRUE, digits = 3)

magma_gs$NAME<-paste(toupper(substr(magma_gs$NAME, 1, 1)), substr(magma_gs$NAME, 2, nchar(magma_gs$NAME)), sep="")

magma_gs<-magma_gs[,c("NAME","NGENES","P","P.FDR") , with=F]
names(magma_gs)<-c('Name','N Genes','P','P.FDR')

magma_gs %>% 
  kable(booktabs = T) %>%
  kable_styling()

```

</details>

<details><summary>Show ATC Enrichment Results</summary>

```{r, eval=T, echo=F}
magma_gs_atc<-fread(paste0('results/',params$gwas,'/magma/magma_drug_targetor_atc_res.csv'))
magma_gs_atc<-magma_gs_atc[order(magma_gs_atc$P),]
magma_gs_atc$P.FDR<-p.adjust(magma_gs_atc$P, method = 'fdr')
  
if(sum(magma_gs_atc$P.FDR < 0.05) >= 10){
  magma_gs_atc<-magma_gs_atc[magma_gs_atc$P.FDR < 0.05,]
} else {
  magma_gs_atc<-magma_gs_atc[1:10,]
}

magma_gs_atc$P<-format(magma_gs_atc$P, scientific = TRUE, digits = 3)
magma_gs_atc$P.FDR<-format(magma_gs_atc$P.FDR, scientific = TRUE, digits = 3)

magma_gs_atc<-magma_gs_atc[,c("ATC","N","P","P.FDR","Name"), with=F]
names(magma_gs_atc)<-c("ATC","N Genes","P","P.FDR","Description")

magma_gs_atc %>% 
  kable(booktabs = T) %>%
  kable_styling()

```

</details>

***

## TWAS-CMAP: Method Comparison

<details><summary>Show Drug Enrichment Comparison</summary>

```{r, eval=T, echo=F}
so_cmap<-fread(paste0('results/',params$gwas,'/twas/cmap/So_res_clean.csv'))
gsea_cmap<-fread(paste0('results/',params$gwas,'/twas/cmap/twas_gsea_cmap.competitive.clean.csv'))

both_cmap<-merge(so_cmap, gsea_cmap[,c('sig_id','Estimate','SE','P','P.CORR'),with=F], by='sig_id')

ggplot(both_cmap, aes(x=-log10(both_cmap$p), y=-log10(both_cmap$P))) +
  geom_point() +
  theme_half_open() +
  geom_abline(intercept =0 , slope = 1) +
  coord_fixed() +
  labs(x="-log10(P)(Average Ranks)", y='-log10(P)(TWAS-GSEA)')


```

</details>

<details><summary>Show ATC Enrichment Comparison</summary>

```{r, eval=T, echo=F}
so_cmap_atc<-fread(paste0('results/',params$gwas,'/twas/cmap/So_res_atc_res.csv'))
gsea_cmap_atc<-fread(paste0('results/',params$gwas,'/twas/cmap/twas_gsea_res_atc_res.csv'))

both_cmap_atc<-merge(so_cmap_atc, gsea_cmap_atc, by='ATC')

ggplot(both_cmap_atc, aes(x=-log10(both_cmap_atc$P.x), y=-log10(both_cmap_atc$P.y))) +
  geom_point() +
  theme_half_open() +
  geom_abline(intercept =0 , slope = 1) +
  coord_fixed() +
  labs(x="-log10(P)(Average Ranks)", y='-log10(P)(TWAS-GSEA)')

```

</details>

***


