---
title: "Drug-targetor ATC enrichment: positive-control validation"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(knitr)
```

***

# Purpose

The pipeline reports enrichment of ATC drug classes using five methods. Three are TWAS-based
(directional, per eQTL panel):

* **Gene-level ("option C")** — each ATC class is a single signed gene set fed to TWAS-GSEA
  (`drug_targetor_atc_genelevel`).
* **Drug-level GLS** — DRUGSETS-style GLS of per-drug `T` on an ATC-class indicator, weighted by the
  drug–drug correlation (`drug_targetor_atc_gls`).
* **Legacy Wilcoxon** — the original per-drug rank-sum test (known to be anti-conservative because
  drugs in a class share target genes; retained only as a cross-check).

Two are MAGMA-based (tissue-independent, non-directional — one result per GWAS):

* **MAGMA Wilcoxon** — MAGMA gene-level association (GWAS + 1000G reference) → per-drug competitive
  gene-set enrichment → a one-sided Wilcoxon of in-class vs out-of-class drug p-values per ATC L3 class
  (`magma_drugtargetor`; the Gaspar & Breen 2017 drug-set approach).
* **MAGMA GLS (DRUGSETS)** — the same MAGMA per-drug enrichment aggregated with the DRUGSETS GLS (Bell,
  de Leeuw, Posthuma): regress the per-drug `BETA/SE` on an ATC-class indicator + size covariates,
  weighted by a drug–drug correlation Σ reconstructed from the MAGMA gene-gene correlations in
  `.genes.raw` (`magma_drugtargetor_gls`). This is the correlation-aware analogue of the MAGMA
  Wilcoxon — the honest test of the same question.

## Consistency of the MAGMA per-drug step with DRUGSETS

GenoDisc's `magma_drug_targetor` runs `magma --gene-results …genes.raw --set-annot …_targetor.gmt`
with no `--model` — i.e. MAGMA's **default competitive, one-sided** gene-set test. This matches
DRUGSETS' per-drug step, with two differences: (1) DRUGSETS adds `--model condition=druggable` (each
drug's enrichment is measured against the druggable-gene background); GenoDisc does not. (2) The
existing MAGMA aggregation is a Wilcoxon on `rank(P)`, whereas DRUGSETS aggregates the per-drug
`BETA/SE` with a GLS. The MAGMA GLS added here reuses the existing (un-conditioned) per-drug betas
(`stat = BETA/SE`) and reconstructs Σ from `.genes.raw` **without** the druggable term, to stay
coherent with the un-conditioned per-drug run — a documented deviation from DRUGSETS (the fully-faithful
`condition=druggable` variant is future work and could recover some signal). (Aside: the existing MAGMA
Wilcoxon script labels the 4-char level "atc_res" and 5-char "level4", offset by one vs the L2=3/L3=4/
L4=5 convention used elsewhere; cosmetic — the scored MAGMA block is 4-char = L3.)

An earlier CAD run (COAD01) produced no clean, directionally-interpretable hit, so before wider use we
asked a simpler question: **for diseases with an unambiguous treating drug class, does that class come
out enriched — and in a trustworthy direction?**

This document records the full analysis: preparing the input GWAS, running the pipeline, and
summarising the results. It is a record of the commands used (chunks are not evaluated at knit time).

# Positive-control design

Six European GWAS were taken from the reference set
(`pipeline_resources/data/secondary_gwas/`), each paired with its canonical treating ATC class and
run on disease-relevant GTEx tissue panel(s) plus `Whole_Blood` as a common comparator.

| Disease | GWAS code | Expected class (ATC level) | Tissue panel(s) + Whole_Blood |
|---|---|---|---|
| Schizophrenia | SCHI04 | `N05A` antipsychotics (L3) | Brain_Cortex, Brain_Frontal_Cortex_BA9 |
| Bipolar disorder | BIPO04 | `N05A` (L3); `N03A` mood stabilisers (L3); `N05AN` lithium (L4) | Brain_Cortex |
| Depression (symptoms) | DEPR02 | `N06A` antidepressants (L3) | Brain_Cortex |
| ADHD | ADHD06 | `N06B` psychostimulants (L3); `N06BA` (L4) | Brain_Cortex |
| Type 2 diabetes | DIAB05 | `A10` (L2); `A10B` oral antidiabetics (L3) | Pancreas, Liver |
| Asthma | ASTH02 | `R03` airway drugs (L2); `R03A` (L3) | Lung |

ATC-level convention in the pipeline: **L2 = 3-char** (`N05`, `A10`), **L3 = 4-char** (`N05A`,
`A10B`), **L4 = 5-char** (`N05AN`, `N06BA`).

***

# Step 1 — Prepare the input GWAS

The reference GWAS are **LDSC-munged** (`CHR SNP BP A1 A2 Z N`; no `BETA`/`SE`/`P`/`FREQ`). The
GenoDisc sumstat cleaner can still ingest them: `insert_beta()` derives `BETA`/`SE` from
`Z + reference MAF + N` (reference MAF is filled during 1KG harmonisation), and the MAF filters skip
gracefully when `FREQ` is absent. The only required column the munged files lack is `P`, which is
recovered from `Z`.

```{r}
# convert_munged.R — add P = 2*pnorm(-|Z|); keep CHR SNP BP A1 A2 Z N P.
library(data.table)

convert_munged <- function(in_gz, out_gz){
  d <- fread(in_gz)
  stopifnot(all(c("SNP","A1","A2","Z","N") %in% names(d)))
  d$P <- 2 * pnorm(-abs(d$Z))
  keep <- intersect(c("CHR","SNP","BP","A1","A2","Z","N","P"), names(d))
  fwrite(d[, ..keep], out_gz, sep = "\t")
}
```

```{r}
# Trait -> label, tissue panels (Whole_Blood appended to every job).
GD <- "/scratch/prj/neurohackpain/GenoDisc"
SG <- file.path(GD, "pipeline_resources/data/secondary_gwas")

traits <- list(
  SCHI04 = list(label = "Schizophrenia",   panels = c("Brain_Cortex","Brain_Frontal_Cortex_BA9","Whole_Blood")),
  BIPO04 = list(label = "Bipolar disorder", panels = c("Brain_Cortex","Whole_Blood")),
  DEPR02 = list(label = "Depressive symptoms", panels = c("Brain_Cortex","Whole_Blood")),
  ADHD06 = list(label = "ADHD",             panels = c("Brain_Cortex","Whole_Blood")),
  DIAB05 = list(label = "Type 2 diabetes",  panels = c("Pancreas","Liver","Whole_Blood")),
  ASTH02 = list(label = "Asthma",           panels = c("Lung","Whole_Blood"))
)
```

# Step 2 — Build gwas_list and config per trait

Each trait gets its own job directory. The config is minimal — only sumstat QC, FUSION TWAS on the
chosen panels, and the drug-targetor / ATC analyses are enabled (no LDSC, cmap, pathway, etc.), to
keep each job lean. Binary/continuous distinction is irrelevant here (no liability-scale h2 step
runs), so `sampling`/`prevalence`/`mean`/`sd` are `NA`.

```{r}
library(data.table)

for(code in names(traits)){
  tr  <- traits[[code]]
  ts  <- format(Sys.time(), "%Y%m%d_%H%M%S")
  job <- file.path(GD, "jobs", sprintf("claude_poscontrol_%s_%s", code, ts))
  dir.create(file.path(job, "input"), recursive = TRUE, showWarnings = FALSE)

  # Convert the munged sumstats into the job's input dir
  convert_munged(file.path(SG, paste0(code, ".sumstats.gz")),
                 file.path(job, "input", paste0(code, ".gz")))

  # gwas_list (space-delimited)
  gl <- data.table(name = code,
                   path = file.path(job, "input", paste0(code, ".gz")),
                   population = "EUR", n = NA, sampling = NA, prevalence = NA,
                   mean = NA, sd = NA, label = tr$label)
  write.table(gl, file.path(job, "gwas_list.txt"), row.names = FALSE, quote = FALSE)

  # config.yaml
  panels_yaml <- paste0("['", paste(tr$panels, collapse = "', '"), "']")
  cfg <- c(
    paste0("outdir: ", job),
    paste0("config_file: ", file.path(job, "config.yaml")),
    paste0("gwas_list: ", file.path(job, "gwas_list.txt")),
    paste0("resdir: ", file.path(GD, "pipeline_resources")),
    "twas_panel_fusion: T",
    paste0("gtex_weights: ", panels_yaml),
    "twas_gsea_drugtargetor: T",
    "twas_gsea_drugtargetor_nondirectional: T",
    "drug_targetor_atc_genelevel: T",
    "drug_targetor_atc_gls: T",
    "magma_drugtargetor: T",         # MAGMA drug-set ATC Wilcoxon (tissue-independent)
    "magma_drugtargetor_gls: T"      # MAGMA DRUGSETS-style drug-level GLS ATC enrichment
  )
  writeLines(cfg, file.path(job, "config.yaml"))
}
```

# Step 3 — Run the pipeline

Each job was submitted exactly as the web service submits jobs — via the launcher, targeting
`results/bundle.tar.gz`, on the dev pipeline. (Locally the equivalent target is the snakemake
`run_package_results` rule.)

```{bash}
GD=/scratch/prj/neurohackpain/GenoDisc

for JOB in $GD/jobs/claude_poscontrol_*; do
  JOBID=$(basename "$JOB")
  CODE=$(echo "$JOBID" | sed -E 's/claude_poscontrol_([A-Z0-9]+)_.*/\1/')
  export GENODISC_PARTITIONS='neurohack_cpu,cpu,interruptible_cpu'
  sbatch --partition=neurohack_cpu,cpu,interruptible_cpu --requeue \
         --cpus-per-task=1 --time=24:00:00 --mem=4G \
         --chdir="$JOB" --job-name="GD-$CODE" \
         --output="$JOB/orchestrator_%j.log" \
         $GD/bin/run_genodisc.sh "$JOB" dev
done

# Equivalent local invocation (per job):
#   source $GD/bin/activate_genodisc.sh
#   cd $GD/repo/current/pipeline
#   snakemake --configfile $GD/repo/current/pipeline/config.yaml <JOB>/config.yaml \
#             --profile profiles/slurm --use-conda \
#             --conda-prefix $GD/conda-envs <JOB>/results/bundle.tar.gz
```

Each job produced `results/bundle.tar.gz` (and an unpacked `results/package/`) containing the ATC
blocks `tx/atc` (legacy Wilcoxon + MAGMA Wilcoxon under `tx/atc$magma`), `tx/atc_genelevel` (option C),
`tx/atc_gls` (TWAS-GSEA GLS).

The **MAGMA GLS** result is produced by two new pipeline scripts (gated by `magma_drugtargetor_gls`,
tissue-independent so one result per GWAS; eval-only — written as CSVs, not yet packaged into a bundle
block): `scripts/compute_magma_drugcorr.R` reconstructs the drug–drug Σ from
`magma_gene_level.genes.raw` (per-chromosome gene-gene-correlation eigen-whitening, eigenvalue prune
0.1, membership residualised on gene covariates → `Σ = cov2cor(VᵀV)`), and
`scripts/format_magma_drugtargetor_gls_results.R` runs the DRUGSETS GLS (`y = BETA/SE`, `X =
[1, class, size, log(size)]`, one-sided upper-tail t) at L2/L3/L4, writing
`results/<GWAS>/magma/magma_drug_targetor_gls_<lv>_res.csv`.

# Step 4 — Score the results

For each (trait, method, panel) we locate the expected class at its ATC level, then record its pooled
`FDR`, direction, and its **rank** by p-value among all classes at that level in that panel. The
tissue-matched panel is the primary read; `Whole_Blood` is a comparator. A "pass" is
`FDR < 0.05` in **any** direction (direction is reported separately as an interpretability check).

```{r}
library(data.table)

# Expected classes per trait, and the matched-tissue keyword.
EXP <- list(
  SCHI04 = list(tissue="Cortex",   classes=list(c("N05A","L3"))),
  BIPO04 = list(tissue="Cortex",   classes=list(c("N05A","L3"),c("N03A","L3"),c("N05AN","L4"))),
  DEPR02 = list(tissue="Cortex",   classes=list(c("N06A","L3"))),
  ADHD06 = list(tissue="Cortex",   classes=list(c("N06BA","L4"),c("N06B","L3"))),
  DIAB05 = list(tissue="Pancreas", classes=list(c("A10","L2"),c("A10B","L3"))),
  ASTH02 = list(tissue="Lung",     classes=list(c("R03","L2"),c("R03A","L3")))
)

read_block <- function(path, slot=NULL){
  if(!file.exists(path)) return(NULL)
  x <- readRDS(path)
  if(!is.null(slot) && is.list(x) && slot %in% names(x)) x <- x[[slot]]
  dt <- as.data.table(x)
  if("ATC Code" %in% names(dt)) setnames(dt, "ATC Code", "code")
  dt
}

# Rank the expected class by P within (panel, level) of a method table.
score_one <- function(dt, code, level, panel_kw, has_level=TRUE){
  if(is.null(dt) || !nrow(dt)) return(NULL)
  d <- dt[grepl(panel_kw, Panel, ignore.case=TRUE)]
  if(has_level && "Level" %in% names(d)) d <- d[Level == level]
  if(!nrow(d)) return(NULL)
  d <- d[order(P)][, rnk := seq_len(.N)]
  idx <- which(d$code == code)                 # explicit column vs arg (avoid data.table scoping)
  if(length(idx) == 0) return(data.table(found=FALSE))
  row <- d[idx[1]]
  data.table(found=TRUE, panel=row$Panel,
             FDR=if("P.FDR" %in% names(row)) signif(row$`P.FDR`,3) else NA_real_,
             Direction=row$Direction, rank=row$rnk, n_at_level=nrow(d),
             percentile=round(row$rnk/nrow(d),3))
}

# MAGMA is tissue-independent + non-directional: score its canonical L3 class per trait.
MAGMA_CLS <- list(SCHI04="N05A", BIPO04="N05A", DEPR02="N06A",
                  ADHD06="N06B", DIAB05="A10B", ASTH02="R03A")

jobdirs <- list.files(file.path(GD, "jobs"), pattern="^claude_poscontrol_", full.names=TRUE)
out <- list()
for(job in jobdirs){
  code <- sub(".*claude_poscontrol_([A-Z0-9]+)_.*", "\\1", basename(job))
  if(!code %in% names(EXP)) next
  tx  <- file.path(job, "results", "package", "gwas", code, "tx")
  gl  <- read_block(file.path(tx, "atc_genelevel.rds"), "twas_gsea")
  gls <- read_block(file.path(tx, "atc_gls.rds"))
  lg  <- read_block(file.path(tx, "atc.rds"), "twas_gsea")
  for(cl in EXP[[code]]$classes){
    for(pt in c("matched","blood")){
      kw <- if(pt=="matched") EXP[[code]]$tissue else "Blood"
      for(m in c("genelevel","gls","legacy")){
        tbl <- switch(m, genelevel=gl, gls=gls, legacy=lg)
        r <- score_one(tbl, cl[1], cl[2], kw, has_level = (m != "legacy"))
        if(is.null(r) || isFALSE(r$found)) next
        r[, `:=`(trait=code, expected=cl[1], level=cl[2], method=m, panel_type=pt)]
        out[[length(out)+1]] <- r
      }
    }
  }
  # MAGMA Wilcoxon: single genome-wide, non-directional table (tx/atc$magma), L3 classes only.
  mg <- read_block(file.path(tx, "atc.rds"), "magma")
  ec <- MAGMA_CLS[[code]]
  if(!is.null(mg) && !is.null(ec)){
    mg <- mg[order(P)][, rnk := seq_len(.N)]
    idx <- which(mg$code == ec)
    if(length(idx)){
      row <- mg[idx[1]]
      out[[length(out)+1]] <- data.table(found=TRUE, panel="(genome-wide)",
        FDR=signif(row$`P.FDR`,3), Direction=NA_character_, rank=row$rnk,
        n_at_level=nrow(mg), percentile=round(row$rnk/nrow(mg),3),
        trait=code, expected=ec, level="L3", method="magma", panel_type="matched")
    }
  }
  # MAGMA GLS (DRUGSETS): read the L3 CSV from the job dir (eval-only, not packaged).
  gf <- file.path(job, "results", code, "magma", "magma_drug_targetor_gls_l3_res.csv")
  if(file.exists(gf) && !is.null(ec)){
    mgls <- fread(gf); setnames(mgls, "Code", "code")
    mgls <- mgls[order(P)][, rnk := seq_len(.N)]
    idx <- which(mgls$code == ec)
    if(length(idx)){
      row <- mgls[idx[1]]
      out[[length(out)+1]] <- data.table(found=TRUE, panel="(genome-wide)",
        FDR=signif(row$FDR,3), Direction=NA_character_, rank=row$rnk,
        n_at_level=nrow(mgls), percentile=round(row$rnk/nrow(mgls),3),
        trait=code, expected=ec, level="L3", method="magma_gls", panel_type="matched")
    }
  }
}
scorecard <- rbindlist(out, fill=TRUE)
```

***

# Results

Canonical treating class per trait, **tissue-matched panel** (MAGMA is genome-wide). Each cell is
`FDR` · rank of the expected class among all classes at its level · direction (— = MAGMA is
non-directional). Significant cells (`FDR < 0.05`) in **bold**.

| Disease | Expected | Gene-level (C) | Drug GLS | Legacy Wilcox | MAGMA Wilcox | MAGMA GLS |
|---|---|---|---|---|---|---|
| Schizophrenia | `N05A` | 0.97 · 229/338 | 0.53 · 37/191 | **0.0011 · 1/145** | **5.6e-8 · 1/111** | 0.70 · 23/126 |
| Type 2 diabetes | `A10B` | 0.85 · top 32% | 0.45 · top 13% | 0.42 · top 23% | **0.0047 · 3/111** | 0.70 · 3/126 |
| ADHD | `N06B` | 0.97 · top 40% | 0.77 · top 48% | 0.82 · top 58% | **0.036 · 6/111** | 0.20 · **1/126** |
| Depression | `N06A` | 0.92 · top 31% | 0.95 · top 60% | 0.87 · top 60% | 0.80 · 30/111 | 1.0 · 101/126 |
| Bipolar | `N05A` | 0.97 · top 75% | 0.74 · top 52% | 0.65 · top 23% | 0.85 · 55/111 | 0.81 · 25/126 |
| Asthma | `R03`* | 0.76 · top 15% | 0.47 · top 19% | 0.99 · top 98% | 0.99 · 62/111 | 0.84 · 63/126 |

Each cell is `FDR` · rank of the expected class among all classes at its level; `FDR < 0.05` in
**bold**. Directional labels (TWAS-GSEA methods) are omitted here for width — they are inconsistent
across methods/tissues (see interpretation); the MAGMA methods are non-directional. \* Legacy and both
MAGMA methods are scored at `R03A` (L3). Rows ordered by MAGMA-Wilcoxon significance.

**Headline counts (of 6 controls):**

| Method | FDR < 0.05 | top 10% by p | direction usable? |
|---|---|---|---|
| **MAGMA Wilcoxon** | **3** (SCZ, T2D, ADHD) | **3** | no (non-directional) |
| MAGMA GLS (DRUGSETS) | 0 | 2 (ADHD rank 1, T2D rank 3) | no (non-directional) |
| Legacy Wilcoxon | 1 (SCZ) | 1 | direction not trustworthy |
| Drug-level GLS | 0 | 1 (T2D) | direction not trustworthy |
| Gene-level (option C) | 0 | 0 | direction not trustworthy |

# Interpretation

* **The MAGMA Wilcoxon's headline results are largely pseudo-replication.** Its three FDR hits look
  impressive, but the Wilcoxon treats the drugs in a class as independent. The correlation-aware
  **MAGMA GLS (DRUGSETS)** — the honest version of the same question — drops all three below FDR 0.05.
  The clearest case is schizophrenia → antipsychotics: `N05A` collapses from rank 1 / FDR 5.6e-8 to
  rank 23 / P 0.13, because its 48 antipsychotics share targets (mean pairwise Σ ≈ 0.19, ~5 effective
  independent drugs). An OLS check reproduces the Wilcoxon (T ≈ 6.5); the Σ-weighted GLS shrinks it to
  T ≈ 1.1 — the correction is doing exactly what it should.
* **But the MAGMA GLS still ranks the right class at the top for the well-behaved controls** — ADHD →
  psychostimulants is rank **1 of 126** (P 0.0016) and T2D → oral antidiabetics rank **3** (P 0.021),
  both nominally significant and top-decile, just not surviving FDR at n = 6. These classes have fewer,
  less-redundant drugs, so they hold up under correction while the antipsychotic cluster does not.
* **All the statistically valid methods now agree.** TWAS option-C, TWAS GLS and MAGMA GLS each recover
  **0/6** at FDR < 0.05: once drug–drug correlation is accounted for, drug-class enrichment from these
  GWAS is at best nominal. The strong signals came only from the two uncorrected Wilcoxon tests.
* **Direction (TWAS-GSEA only) remains untrustworthy** — the "Opposes/Matches" label flips across
  methods and tissues (COAD01 diagnosis). The MAGMA methods sidestep this (non-directional magnitude).
* **Consistency caveat / next step:** the MAGMA GLS here reuses the un-conditioned per-drug betas and a
  Σ without the druggable term (see the consistency section). The fully-faithful DRUGSETS variant
  (`--model condition=druggable`) removes the common druggable-gene component and could lift some
  classes back toward significance — worth testing before drawing a final line under the drug-level
  approach.

# Recommendation

* **Use the MAGMA GLS (DRUGSETS), not the MAGMA Wilcoxon, as the drug-level ATC readout.** The
  Wilcoxon's strong hits are inflated by drug–drug correlation; the GLS is the correlation-aware, honest
  version of the same test. Report it as a non-directional competitive enrichment (with the coefficient
  and one-sided p), and read the nominal ranking, not FDR-significance, as the signal.
* **Be candid that, once correlation is accounted for, no drug class survives FDR across these six
  controls** — every valid method (TWAS option-C, TWAS GLS, MAGMA GLS) agrees. The honest message is
  "top-ranked, nominally-enriched" (e.g. ADHD → psychostimulants, T2D → antidiabetics), not "validated".
* **Do not present the TWAS-GSEA ATC methods as a validated repurposing signal** and do not surface the
  directional "reverses disease" claim — the controls don't support it.
* **Before finalising, test the fully-faithful DRUGSETS variant** (`--model condition=druggable` per-drug
  run + druggable term in Σ). If it lifts the well-behaved controls (ADHD, T2D) to FDR-significance, it
  becomes the recommended default; if not, the drug-level approach is confirmed underpowered at these
  GWAS sizes.

# Caveats

* The reference GWAS are HapMap3-only (LDSC-munged), which modestly reduces TWAS gene coverage
  relative to full summary statistics (this hurts the TWAS-GSEA methods more than MAGMA).
* Drug-level methods have few drugs per class (e.g. `R03A` ≈ 2, `A10B` ≈ 12 in the matched panel),
  so are inherently low-powered.
* MAGMA is scored at ATC L3 only (the packaged level) and is non-directional, so it is a comparator,
  not a like-for-like replacement for the directional TWAS-GSEA rows.
* A "miss" can reflect GWAS power, tissue coverage, or drug-annotation gaps rather than a broken
  method; absence of signal is not proof of no effect.

***
