---
title: "Drug-targetor ATC enrichment: positive-control validation"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(knitr)
```

***

# Purpose

The pipeline reports enrichment of ATC drug classes among a GWAS's TWAS signal using three
methods:

* **Gene-level ("option C")** — each ATC class is a single signed gene set fed to TWAS-GSEA
  (`drug_targetor_atc_genelevel`).
* **Drug-level GLS** — DRUGSETS-style GLS of per-drug `T` on an ATC-class indicator, weighted by the
  drug–drug correlation (`drug_targetor_atc_gls`).
* **Legacy Wilcoxon** — the original per-drug rank-sum test (known to be anti-conservative because
  drugs in a class share target genes; retained only as a cross-check).

An earlier CAD run (COAD01) produced no clean, directionally-interpretable hit, so before wider use we
asked a simpler question: **for diseases with an unambiguous treating drug class, does that class come
out enriched — and in a trustworthy direction?**

This document records the full analysis: preparing the input GWAS, running the pipeline, and
summarising the results. It is a record of the commands used (chunks are not evaluated at knit time).

# Positive-control design

Six European GWAS were taken from the reference set
(`pipeline_resources/data/secondary_gwas/`), each paired with its canonical treating ATC class and
run on disease-relevant GTEx tissue panel(s) plus `Whole_Blood` as a common comparator.

| Disease | GWAS code | Expected class (ATC level) | Tissue panel(s) + Whole_Blood |
|---|---|---|---|
| Schizophrenia | SCHI04 | `N05A` antipsychotics (L3) | Brain_Cortex, Brain_Frontal_Cortex_BA9 |
| Bipolar disorder | BIPO04 | `N05A` (L3); `N03A` mood stabilisers (L3); `N05AN` lithium (L4) | Brain_Cortex |
| Depression (symptoms) | DEPR02 | `N06A` antidepressants (L3) | Brain_Cortex |
| ADHD | ADHD06 | `N06B` psychostimulants (L3); `N06BA` (L4) | Brain_Cortex |
| Type 2 diabetes | DIAB05 | `A10` (L2); `A10B` oral antidiabetics (L3) | Pancreas, Liver |
| Asthma | ASTH02 | `R03` airway drugs (L2); `R03A` (L3) | Lung |

ATC-level convention in the pipeline: **L2 = 3-char** (`N05`, `A10`), **L3 = 4-char** (`N05A`,
`A10B`), **L4 = 5-char** (`N05AN`, `N06BA`).

***

# Step 1 — Prepare the input GWAS

The reference GWAS are **LDSC-munged** (`CHR SNP BP A1 A2 Z N`; no `BETA`/`SE`/`P`/`FREQ`). The
GenoDisc sumstat cleaner can still ingest them: `insert_beta()` derives `BETA`/`SE` from
`Z + reference MAF + N` (reference MAF is filled during 1KG harmonisation), and the MAF filters skip
gracefully when `FREQ` is absent. The only required column the munged files lack is `P`, which is
recovered from `Z`.

```{r}
# convert_munged.R — add P = 2*pnorm(-|Z|); keep CHR SNP BP A1 A2 Z N P.
library(data.table)

convert_munged <- function(in_gz, out_gz){
  d <- fread(in_gz)
  stopifnot(all(c("SNP","A1","A2","Z","N") %in% names(d)))
  d$P <- 2 * pnorm(-abs(d$Z))
  keep <- intersect(c("CHR","SNP","BP","A1","A2","Z","N","P"), names(d))
  fwrite(d[, ..keep], out_gz, sep = "\t")
}
```

```{r}
# Trait -> label, tissue panels (Whole_Blood appended to every job).
GD <- "/scratch/prj/neurohackpain/GenoDisc"
SG <- file.path(GD, "pipeline_resources/data/secondary_gwas")

traits <- list(
  SCHI04 = list(label = "Schizophrenia",   panels = c("Brain_Cortex","Brain_Frontal_Cortex_BA9","Whole_Blood")),
  BIPO04 = list(label = "Bipolar disorder", panels = c("Brain_Cortex","Whole_Blood")),
  DEPR02 = list(label = "Depressive symptoms", panels = c("Brain_Cortex","Whole_Blood")),
  ADHD06 = list(label = "ADHD",             panels = c("Brain_Cortex","Whole_Blood")),
  DIAB05 = list(label = "Type 2 diabetes",  panels = c("Pancreas","Liver","Whole_Blood")),
  ASTH02 = list(label = "Asthma",           panels = c("Lung","Whole_Blood"))
)
```

# Step 2 — Build gwas_list and config per trait

Each trait gets its own job directory. The config is minimal — only sumstat QC, FUSION TWAS on the
chosen panels, and the drug-targetor / ATC analyses are enabled (no LDSC, cmap, pathway, etc.), to
keep each job lean. Binary/continuous distinction is irrelevant here (no liability-scale h2 step
runs), so `sampling`/`prevalence`/`mean`/`sd` are `NA`.

```{r}
library(data.table)

for(code in names(traits)){
  tr  <- traits[[code]]
  ts  <- format(Sys.time(), "%Y%m%d_%H%M%S")
  job <- file.path(GD, "jobs", sprintf("claude_poscontrol_%s_%s", code, ts))
  dir.create(file.path(job, "input"), recursive = TRUE, showWarnings = FALSE)

  # Convert the munged sumstats into the job's input dir
  convert_munged(file.path(SG, paste0(code, ".sumstats.gz")),
                 file.path(job, "input", paste0(code, ".gz")))

  # gwas_list (space-delimited)
  gl <- data.table(name = code,
                   path = file.path(job, "input", paste0(code, ".gz")),
                   population = "EUR", n = NA, sampling = NA, prevalence = NA,
                   mean = NA, sd = NA, label = tr$label)
  write.table(gl, file.path(job, "gwas_list.txt"), row.names = FALSE, quote = FALSE)

  # config.yaml
  panels_yaml <- paste0("['", paste(tr$panels, collapse = "', '"), "']")
  cfg <- c(
    paste0("outdir: ", job),
    paste0("config_file: ", file.path(job, "config.yaml")),
    paste0("gwas_list: ", file.path(job, "gwas_list.txt")),
    paste0("resdir: ", file.path(GD, "pipeline_resources")),
    "twas_panel_fusion: T",
    paste0("gtex_weights: ", panels_yaml),
    "twas_gsea_drugtargetor: T",
    "twas_gsea_drugtargetor_nondirectional: T",
    "drug_targetor_atc_genelevel: T",
    "drug_targetor_atc_gls: T"
  )
  writeLines(cfg, file.path(job, "config.yaml"))
}
```

# Step 3 — Run the pipeline

Each job was submitted exactly as the web service submits jobs — via the launcher, targeting
`results/bundle.tar.gz`, on the dev pipeline. (Locally the equivalent target is the snakemake
`run_package_results` rule.)

```{bash}
GD=/scratch/prj/neurohackpain/GenoDisc

for JOB in $GD/jobs/claude_poscontrol_*; do
  JOBID=$(basename "$JOB")
  CODE=$(echo "$JOBID" | sed -E 's/claude_poscontrol_([A-Z0-9]+)_.*/\1/')
  export GENODISC_PARTITIONS='neurohack_cpu,cpu,interruptible_cpu'
  sbatch --partition=neurohack_cpu,cpu,interruptible_cpu --requeue \
         --cpus-per-task=1 --time=24:00:00 --mem=4G \
         --chdir="$JOB" --job-name="GD-$CODE" \
         --output="$JOB/orchestrator_%j.log" \
         $GD/bin/run_genodisc.sh "$JOB" dev
done

# Equivalent local invocation (per job):
#   source $GD/bin/activate_genodisc.sh
#   cd $GD/repo/current/pipeline
#   snakemake --configfile $GD/repo/current/pipeline/config.yaml <JOB>/config.yaml \
#             --profile profiles/slurm --use-conda \
#             --conda-prefix $GD/conda-envs <JOB>/results/bundle.tar.gz
```

Each job produced `results/bundle.tar.gz` (and an unpacked `results/package/`) containing the ATC
blocks `tx/atc` (legacy), `tx/atc_genelevel` (option C), `tx/atc_gls` (GLS).

# Step 4 — Score the results

For each (trait, method, panel) we locate the expected class at its ATC level, then record its pooled
`FDR`, direction, and its **rank** by p-value among all classes at that level in that panel. The
tissue-matched panel is the primary read; `Whole_Blood` is a comparator. A "pass" is
`FDR < 0.05` in **any** direction (direction is reported separately as an interpretability check).

```{r}
library(data.table)

# Expected classes per trait, and the matched-tissue keyword.
EXP <- list(
  SCHI04 = list(tissue="Cortex",   classes=list(c("N05A","L3"))),
  BIPO04 = list(tissue="Cortex",   classes=list(c("N05A","L3"),c("N03A","L3"),c("N05AN","L4"))),
  DEPR02 = list(tissue="Cortex",   classes=list(c("N06A","L3"))),
  ADHD06 = list(tissue="Cortex",   classes=list(c("N06BA","L4"),c("N06B","L3"))),
  DIAB05 = list(tissue="Pancreas", classes=list(c("A10","L2"),c("A10B","L3"))),
  ASTH02 = list(tissue="Lung",     classes=list(c("R03","L2"),c("R03A","L3")))
)

read_block <- function(path, slot=NULL){
  if(!file.exists(path)) return(NULL)
  x <- readRDS(path)
  if(!is.null(slot) && is.list(x) && slot %in% names(x)) x <- x[[slot]]
  dt <- as.data.table(x)
  if("ATC Code" %in% names(dt)) setnames(dt, "ATC Code", "code")
  dt
}

# Rank the expected class by P within (panel, level) of a method table.
score_one <- function(dt, code, level, panel_kw, has_level=TRUE){
  if(is.null(dt) || !nrow(dt)) return(NULL)
  d <- dt[grepl(panel_kw, Panel, ignore.case=TRUE)]
  if(has_level && "Level" %in% names(d)) d <- d[Level == level]
  if(!nrow(d)) return(NULL)
  d <- d[order(P)][, rnk := seq_len(.N)]
  idx <- which(d$code == code)                 # explicit column vs arg (avoid data.table scoping)
  if(length(idx) == 0) return(data.table(found=FALSE))
  row <- d[idx[1]]
  data.table(found=TRUE, panel=row$Panel,
             FDR=if("P.FDR" %in% names(row)) signif(row$`P.FDR`,3) else NA_real_,
             Direction=row$Direction, rank=row$rnk, n_at_level=nrow(d),
             percentile=round(row$rnk/nrow(d),3))
}

jobdirs <- list.files(file.path(GD, "jobs"), pattern="^claude_poscontrol_", full.names=TRUE)
out <- list()
for(job in jobdirs){
  code <- sub(".*claude_poscontrol_([A-Z0-9]+)_.*", "\\1", basename(job))
  if(!code %in% names(EXP)) next
  tx  <- file.path(job, "results", "package", "gwas", code, "tx")
  gl  <- read_block(file.path(tx, "atc_genelevel.rds"), "twas_gsea")
  gls <- read_block(file.path(tx, "atc_gls.rds"))
  lg  <- read_block(file.path(tx, "atc.rds"), "twas_gsea")
  for(cl in EXP[[code]]$classes){
    for(pt in c("matched","blood")){
      kw <- if(pt=="matched") EXP[[code]]$tissue else "Blood"
      for(m in c("genelevel","gls","legacy")){
        tbl <- switch(m, genelevel=gl, gls=gls, legacy=lg)
        r <- score_one(tbl, cl[1], cl[2], kw, has_level = (m != "legacy"))
        if(is.null(r) || isFALSE(r$found)) next
        r[, `:=`(trait=code, expected=cl[1], level=cl[2], method=m, panel_type=pt)]
        out[[length(out)+1]] <- r
      }
    }
  }
}
scorecard <- rbindlist(out, fill=TRUE)
```

***

# Results

Canonical treating class per trait, **tissue-matched panel**. Each cell is `FDR` · rank of the
expected class among all classes at its level · direction.

| Disease | Expected | Gene-level (option C) | Drug-level GLS | Legacy Wilcoxon |
|---|---|---|---|---|
| Schizophrenia | `N05A` | 0.97 · 229/338 · Matches | 0.53 · 37/191 · Matches | **0.0011 · 1/145 · Opposes** |
| Bipolar | `N05A` | 0.97 · top 75% · Opposes | 0.74 · top 52% · Opposes | 0.65 · top 23% · Opposes |
| Depression | `N06A` | 0.92 · top 31% · Opposes | 0.95 · top 60% · Opposes | 0.87 · top 60% · Matches |
| ADHD | `N06B` | 0.97 · top 40% · Opposes | 0.77 · top 48% · Matches | 0.82 · top 58% · Opposes |
| Type 2 diabetes | `A10B` | 0.85 · top 32% · Opposes | 0.45 · top 13% · Opposes | 0.42 · top 23% · Opposes |
| Asthma | `R03`* | 0.76 · top 15% · Matches | 0.47 · top 19% · Matches | 0.99 · top 98% · Matches |

\* Legacy has no ATC-level column, so asthma's legacy cell is scored at `R03A`.

**Headline counts (of 6 controls, tissue-matched panel):**

| Method | FDR < 0.05 | top 10% by p | direction = "Opposes" |
|---|---|---|---|
| Legacy Wilcoxon | 1 (schizophrenia) | 1 | 4 |
| Drug-level GLS | 0 | 1 (T2D `A10` L2) | 3 |
| Gene-level (option C) | 0 | 0 | 3 |

# Interpretation

* **Only one of six positive controls is significant** — schizophrenia → antipsychotics — and only
  under the **legacy Wilcoxon** (FDR 0.001, rank 1, correct "Opposes" direction). That test is the
  statistically invalid one; its low p-value is partly the pseudo-replication we set out to remove.
* **Both valid methods recover zero of six** at FDR < 0.05. Removing the drug-correlation inflation
  removes most of the apparent signal. The closest they come is type-2-diabetes → oral antidiabetics
  (GLS, top ~13%, correct direction) — suggestive, not significant.
* **Direction is not trustworthy.** For the same class the "Opposes / Matches disease" label flips
  between blood and the matched tissue and across methods (e.g. schizophrenia → antipsychotics reads
  "Matches" under the valid methods but "Opposes" under legacy). This confirms the earlier COAD01
  diagnosis that direction is largely a gene-coverage artifact.
* Schizophrenia is by far the best-powered GWAS here, and the one case where signal survives — i.e.
  the valid methods may be **correctly calibrated but underpowered** at realistic GWAS sizes rather
  than fundamentally wrong.

# Recommendation

* Do not present ATC enrichment as a validated drug-repurposing signal, and do not surface the
  directional "reverses disease" claim — it is not supported by these controls.
* If retained, frame it as **exploratory and direction-agnostic** ("this class's target genes are
  enriched among your disease-associated genes"), with a power caveat.
* Re-test the valid methods on the very largest available GWAS before shipping; keep the legacy
  Wilcoxon out of any headline claim.

# Caveats

* The reference GWAS are HapMap3-only (LDSC-munged), which modestly reduces TWAS gene coverage
  relative to full summary statistics.
* Drug-level methods have few drugs per class (e.g. `R03A` ≈ 2, `A10B` ≈ 12 in the matched panel),
  so are inherently low-powered.
* A "miss" can reflect GWAS power, tissue coverage, or drug-annotation gaps rather than a broken
  method; absence of signal is not proof of no effect.

***
