---
title: "GenoDisc Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
library(data.table)
```

* Download and format GWAS sumstats for brain related disorders
* Run pipeline for all GWAS
* Collate and compare results

# V1

***

## Download and format GWAS sumstats

***

### Schizophrenia (SCZ)

PMID: 35396580
EUR only

```{r}
dir.create('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/SCZ', recursive = T)
system('wget --no-check-certificate -O /users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/SCZ/SCZ.original.gz https://figshare.com/ndownloader/files/34517828')

ss<-fread('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/SCZ/SCZ.original.gz')

# Update column names. Use NEFF as N
# I am not sure how PGS is computing NEFF but appears wrong
names(ss)<-c('CHR','SNP','ORIGBP','A1','A2','FCAS','FCON','INFO','BETA','SE','P','NCAS','NCON','NEFF')
ss$N<-4/((1/ss$NCAS)+(1/ss$NCON))

# Calculate overall FREQ
ss$FREQ<-((ss$FCAS*ss$NCAS) + (ss$FCON*ss$NCON))/(ss$NCAS + ss$NCON)

# Keep only relevent columns
ss<-ss[,c('SNP','A1','A2','INFO','BETA','SE','P','N','FREQ'), with=F]

fwrite(ss, '/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/SCZ/SCZ.reformat.gz', sep=' ', na='NA', quote=F)

```

***

### Bipolar Disorder (BIP)

PMID: 39843750 (https://pgc.unc.edu/for-researchers/download-results/)

```{r}
dir.create('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/BIP', recursive = T)

ss<-fread('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/BIP/BIP.original.gz')

# Update column names. Use NEFF as N
# I am not sure how PGS is computing NEFF but appears wrong
names(ss)<-c('CHR','ORIGBP','SNP','A1','A2','BETA','SE','P','NGT','FCAS','FCON','INFO','NEFFDIV2','NCAS','NCON','DIR')
ss$N<-4/((1/ss$NCAS)+(1/ss$NCON))

# Calculate overall FREQ
ss$FREQ<-((ss$FCAS*ss$NCAS) + (ss$FCON*ss$NCON))/(ss$NCAS + ss$NCON)

# Keep only relevent columns
ss<-ss[,c('CHR','ORIGBP','SNP','A1','A2','INFO','BETA','SE','P','N','FREQ'), with=F]

fwrite(ss, '/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/BIP/BIP.reformat.gz', sep=' ', na='NA', quote=F)

```

***

### Major Depressive Disorder (MDD)

I have access to the latest sumstat (incl. 23andMe). Not yet published.

```{r}

dir.create('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/MDD', recursive = T)

# Read in the sumstats
ss<-fread('/users/k1806347/brc_scratch/Software/mdd-meta_2/mdd-meta/results/distribution/daner_pgc_mdd_full_eur_hg19_v3.49.24.11.gz')

# Update column names
names(ss)<-c('CHR','SNP','ORIGBP','A1','A2','FCAS','FCON','INFO','OR','SE','P','NGT','DIR','HETISQT','HETDF','HETP','NCAS','NCON','NEFFDIV2')
ss$N<-4/((1/ss$NCAS)+(1/ss$NCON))

# Calculate overall FREQ
ss$FREQ<-((ss$FCAS*ss$NCAS) + (ss$FCON*ss$NCON))/(ss$NCAS + ss$NCON)

# Keep only relevent columns
ss<-ss[,c('SNP','A1','A2','INFO','OR','SE','P','N','FREQ'), with=F]

fwrite(ss, '/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/MDD/MDD.reformat.gz', sep=' ', na='NA', quote=F)

```

***

### ADHD

PMID: 36702997 (https://pgc.unc.edu/for-researchers/download-results/)

```{r}

dir.create('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/ADHD', recursive = T)

# Read in the sumstats
ss<-fread('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/ADHD/ADHD.original.gz')

names(ss) <- c('CHR','SNP','ORIGBP','A1','A2','FCAS','FCON','INFO','OR','SE','P','NGT','Direction','HetISqt','HetDf','HetPVa','NCAS','NCON','NEFF')

# Update to NEFF
ss$N <- 4 / ((1 / ss$NCAS) + (1 / ss$NCON))

# Calculate overall FREQ
ss$FREQ<-((ss$FCAS*ss$NCAS) + (ss$FCON*ss$NCON))/(ss$NCAS + ss$NCON)

ss <- ss[, c('CHR','SNP','ORIGBP','A1','A2','FREQ','INFO','OR','SE','P','N'), with = F]

fwrite(ss, '/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/ADHD/ADHD.reformat.gz', sep=' ', na='NA', quote=F)

```

***

### Autism Spectrum Disorders (ASD)

PMID: 30804558
Already available on Rosalind: /scratch/prj/gwas_sumstats/cleaned/AUTI07.gz

```{r}

dir.create('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/ASD', recursive = T)

# Read in the sumstats
ss<-fread('/scratch/prj/gwas_sumstats/cleaned/AUTI07.gz')

# Update to NEFF
ss$N <- NULL
ss$N <- 4 / ((1 / ss$Ncas) + (1 / ss$Ncon))
ss$Ncas <- NULL
ss$Ncon <- NULL

fwrite(ss, '/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/ASD/ASD.reformat.gz', sep=' ', na='NA', quote=F)

```

***

### Amyotrophic Lateral Sclerosis (ALS)

PMID: 34873335
European only.
Already available on Rosalind: /users/k1806347/brc_scratch/Data/GWAS_sumstats/ALS/GWAS/ALS_sumstats_EUR_only_reformat.txt.gz

***

### Alzheimer's Disease (ALZ)

PMID: 34493870
Excludes 23andMe.

```{r}
dir.create('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/ALZ', recursive = T)
system('wget --no-check-certificate -O /users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/ALZ/ALZ.original.gz https://ctg.cncr.nl/documents/p1651/PGCALZ2sumstatsExcluding23andMe.txt.gz')

ss<-fread('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/ALZ/ALZ.original.gz')

# Update column names.
names(ss)<-c('CHR','ORIGBP','A1','A2','BETA','P','N')

fwrite(ss, '/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/ALZ/ALZ.reformat.gz', sep=' ', na='NA', quote=F)

```


***

### Parkinson's Disease (PRK)

Publicly available version
PMID: 31701892

```{r}
dir.create('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/PRK', recursive = T)
system('wget --load-cookies /tmp/cookies.txt "https://docs.google.com/uc?export=download&confirm=$(wget --quiet --save-cookies /tmp/cookies.txt --keep-session-cookies --no-check-certificate 'https://docs.google.com/uc?export=download&id=1FZ9UL99LAqyWnyNBxxlx6qOUlfAnublN' -O- | sed -rn 's/.*confirm=([0-9A-Za-z_]+).*/\1\n/p')&id=1FZ9UL99LAqyWnyNBxxlx6qOUlfAnublN" -O /users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/PRK/PRK.original.zip && rm -rf /tmp/cookies.txt')
setwd('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/PRK/')
system('unzip /users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/PRK/PRK.original.zip')
system('rm /users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/PRK/PRK.original.zip')

ss<-fread('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/PRK/nallsEtAl2019_excluding23andMe_allVariants.tab')

ss$CHR<-gsub('chr','',gsub(':.*','',ss$SNP))
ss$BP<-gsub('.*:','',ss$SNP)

# Update column names.
names(ss)<-c('SNP','A1','A2','FREQ','BETA','SE','P','NCAS','NCON','CHR','ORIGBP')

ss$N<-4/((1/ss$NCAS)+(1/ss$NCON))

# Keep only relevent columns
ss<-ss[,c('CHR','ORIGBP','A1','A2','BETA','SE','P','N','FREQ'), with=F]

fwrite(ss, '/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/PRK/PRK.reformat.gz', sep=' ', na='NA', quote=F)

```

***

### Migraine (MIG)

PMID: 35115687
Excl. 23andMe
Requested from Dale Nyholt.

```{r}

dir.create('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/MIG', recursive = T)

ss<-fread('/users/k1806347/brc_scratch/Data/GWAS_sumstats/Migraine_Hautakangas_excl23andMe/migraine_IHGC2021_no23andMe_gwama_v2.txt.gz')

ss<-ss[,c('chromosome','position','reference_allele','other_allele','eaf','beta','se','p.value','Neff'), with=F]
names(ss)<-c('CHR','ORIGBP','A1','A2','FREQ','BETA','SE','P','N')

fwrite(ss, '/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/MIG/MIG.reformat.gz', sep=' ', na='NA', quote=F)

```

***

## Create configuration

```{r}

dir.create('/users/k1806347/oliverpainfel/Analyses/genodisc_validation/config/', recursive = T)

library(data.table)

####
# Create gwas_list
####

gwas_files <- list.files('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper', pattern = 'reformat', full.names = T, recursive = T)

gwas_list <- data.table(
  name = gsub('/.*','', gsub('.*GenoDisc_paper/', '', gwas_files)),
  path = gwas_files,
  population = 'EUR',
    n = NA,
    sampling = NA,
    prevalence = NA,
    mean = NA,
    sd = NA,
    label = gsub('/.*','', gsub('.*GenoDisc_paper//', '', gwas_files))
)

write.table(gwas_list, '/users/k1806347/oliverpainfel/Analyses/genodisc_validation/config/gwas_list_neuropsych.txt', row.names = F)

####
# Create config file
####

config <- c(
  "outdir: /users/k1806347/oliverpainfel/Analyses/genodisc_validation/output_neuropsych", 
  "config_file: /users/k1806347/oliverpainfel/Analyses/genodisc_validation/config/config_neuropsych.yaml", 
  "gwas_list: /users/k1806347/oliverpainfel/Analyses/genodisc_validation/config/gwas_list_neuropsych.txt", 
#  "gtex_weights: ['Brain_Amygdala','Brain_Anterior_cingulate_cortex_BA24','Brain_Caudate_basal_ganglia','Brain_Cerebellar_Hemisphere','Brain_Cerebellum','Brain_Cortex','Brain_Frontal_Cortex_BA9','Brain_Hippocampus','Brain_Hypothalamus','Brain_Nucleus_accumbens_basal_ganglia','Brain_Putamen_basal_ganglia','Brain_Spinal_cord_cervical_c-1','Brain_Substantia_nigra','Whole_Blood']", 
  "clump: T", 
  "cojo: T", 
  "finemap: T", 
  "ldsc: T", 
  "magma_gene: T", 
  "magma_drugtargetor: T", 
  "twas_panel_fusion: T", 
  "non_gtex_weights: ['CMC.BRAIN.RNASEQ','CMC.BRAIN.RNASEQ_SPLICING','NTR.BLOOD.RNAARR','YFS.BLOOD.RNAARR']",
  "twas_panel_psychencode: T",
  "twas_conditional: F", 
  "twas_gsea_drugtargetor: T", 
  "smr_expression_panel_eqtlgen: T",
  "tissue_magma: T",
  "rosmap_fusion: /scratch/prj/oliverpainfel/Data/ROSMAP-PWAS/ROSMAP.n376.fusion.WEIGHTS.zip",
  "banner_fusion: /scratch/prj/oliverpainfel/Data/Banner-PWAS/Banner.n152.fusion.WEIGHTS.zip",
  "rosmap_smr: /scratch/prj/oliverpainfel/Data/ROSMAP_pQTL/ROSMAP.n376.pQTL.txt",
  "smr_expression_panel_psychencode: T",
  "smr_expression_panel_metabrain_basalganglia: T",
  "smr_expression_panel_metabrain_cerebellum: T",
  "smr_expression_panel_metabrain_cortex: T",
  "smr_expression_panel_metabrain_hippocampus: T",
  "smr_expression_panel_metabrain_spinalcord: T",
  "pwas_panel_rosmap: T",
  "pwas_panel_banner: T",
  "smr_protein_panel_rosmap: T",
  "twas_gsea_cmap: T",
  "cmap_level5_prop: /users/k1806347/oliverpainfel/Data/CMAP/reprocessed/CMAP_level5/lvl5Allcompounds_population_TWAS-GSEA_prop.rds",
  "cmap_compoundinfo: /users/k1806347/oliverpainfel/Data/CMAP/compoundinfo_beta.txt",
  "cmap_cell_iname: ['HT29','A549','HA1E','A375','VCAP','PC3','MCF7']",
  "cmap_pert_itime: ['24h']",
  "cmap_pert_idose: ['10uM']",
  "twas_gsea_drugtargetor_nondirectional: T",
  "gcsc: T",
  "gcsc_tissues: ['Brain_Amygdala','Brain_Cerebellum','Brain_Cortex','Whole_Blood']"
#  "gcsc_tissues: ['Brain_Amygdala','Brain_Anterior_cingulate_cortex_BA24','Brain_Caudate_basal_ganglia','Brain_Cerebellar_Hemisphere','Brain_Cerebellum','Brain_Cortex','Brain_Frontal_Cortex_BA9','Brain_Hippocampus','Brain_Hypothalamus','Brain_Nucleus_accumbens_basal_ganglia','Brain_Putamen_basal_ganglia','Brain_Spinal_cord_cervical_c-1','Brain_Substantia_nigra','Whole_Blood']"
)


    
write.table(config, '/users/k1806347/oliverpainfel/Analyses/genodisc_validation/config/config_neuropsych.yaml', col.names=F, row.names=F, quote=F)
```

***

## Run pipeline for all GWAS

```{bash, eval = F}
cd /users/k1806347/oliverpainfel/Software/MyGit/GenoDisc/pipeline
conda activate genodisc
  
snakemake \
  --configfile=/users/k1806347/oliverpainfel/Analyses/genodisc_validation/config/config_neuropsych.yaml \
  --use-conda \
  --profile slurm \
  run_package_results \
  -n

```

***

# V2

- Run using latest neurohack version.
- Add rG analysis
- Drop CMAP analysis

# Run using neurohack version

```{r}
####
# Make a gencor_gwas_list with all GWAS in reference
####

secondary_gwas <- fread('/scratch/prj/neurohackpain/GenoDisc/pipeline_resources/data/secondary_gwas/sumstat_repo.csv')
gencor_gwas_list <- data.table(
  name = secondary_gwas$code,
  path = paste0('/scratch/prj/neurohackpain/GenoDisc/pipeline_resources/data/secondary_gwas/', secondary_gwas$code, '.sumstats.gz'),
  population = 'EUR',
  n = NA,
  sampling = NA,
  prevalence = NA,
  mean = NA,
  sd = NA,
  label = paste0('"',secondary_gwas$short_label, '"'),
  trait_category = paste0('"',secondary_gwas$category, '"')
)

write.table(gencor_gwas_list, 
            '/users/k1806347/oliverpainfel/Analyses/genodisc_validation/config/gencor_gwas_list.txt',
            col.names = T, 
            row.names = F, 
            quote = F)

####
# Create config file
####

config <- c(
  "outdir: /users/k1806347/oliverpainfel/Analyses/genodisc_validation/output_neuropsych_v2", 
  "config_file: /users/k1806347/oliverpainfel/Analyses/genodisc_validation/config/config_neuropsych_v2.yaml", 
  "gwas_list: /users/k1806347/oliverpainfel/Analyses/genodisc_validation/config/gwas_list_neuropsych.txt", 
  "magma_gene: T", 
  "magma_drugtargetor: T", 
  "tissue_magma: T",
  "magma_pathway: T",
  "pathway_gmt_dir: /scratch/prj/neurohackpain/GenoDisc/pipeline_resources/data/msigdb",
  "magma_interaction:",
  "  enabled: true",
  "  min_set_size: 100",
  "  fdr_threshold: 0.05",
  "  n_cores: 8",
  "ldsc: T",
  "gencor_gwas_list: /users/k1806347/oliverpainfel/Analyses/genodisc_validation/config/gencor_gwas_list.txt",
  "gencor_within_gwas_list: T",
  "resdir: /scratch/prj/neurohackpain/GenoDisc/pipeline_resources"
)

write.table(config, '/users/k1806347/oliverpainfel/Analyses/genodisc_validation/config/config_neuropsych_v2.yaml', col.names=F, row.names=F, quote=F)

```

```{bash}

# Change resource directory
source /scratch/prj/neurohackpain/GenoDisc/bin/activate_genodisc.sh
cd /scratch/prj/neurohackpain/GenoDisc/repo/current/pipeline

snakemake \
  -n \
  --profile slurm \
  --configfile=/users/k1806347/oliverpainfel/Analyses/genodisc_validation/config/config_neuropsych_v2.yaml \
  --use-conda \
  --conda-prefix /scratch/prj/neurohackpain/GenoDisc/conda-envs \
  run_package_results

```


***

# Get effective sample size of each GWAS

```{r}

library(data.table)

gwas_files <- list.files('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper', pattern = 'reformat', full.names = T, recursive = T)

effective_n <- NULL
for(i in gwas_files){
  dat<-fread(i)
  med_n <- median(dat$N)
  effective_n <- rbind(
    effective_n,
    data.table(
      gwas = gsub('\\/.*','', gsub('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/','', i)),
      n_eff = med_n
    )
  )
}


effective_n <- NULL
for(i in gwas_files){
  dat<-fread(i)
  med_n <- median(dat$N)
  effective_n <- rbind(
    effective_n,
    data.table(
      gwas = gsub('\\/.*','', gsub('/users/k1806347/oliverpainfel/Data/GWAS_sumstats/GenoDisc_paper/','', i)),
      n_eff = med_n
    )
  )
}

/users/k1806347/oliverpainfel/Analyses/genodisc_validation/output_neuropsych

```

***

# MAGMA gene-set x tissue interaction summary

Cross-GWAS summary of the post-hoc MAGMA gene-set x tissue interaction
module (`magma_gene_set_by_tissue_interaction` rule) for the 9 neuro/psych
GWAS. Reads `interaction_results.tsv` and `interaction_calibration.tsv`
per GWAS from `output_neuropsych_v2/results/<gwas>/magma/interaction/`.
No pipeline re-run needed — this whole section reads the finished
outputs.

```{r interaction-load, eval = T}
library(data.table)
library(ggplot2)

outdir_root <- '/users/k1806347/oliverpainfel/Analyses/genodisc_validation/output_neuropsych_v2/results'
gwas_list   <- c('SCZ','ADHD','ALS','ALZ','ASD','BIP','MDD','MIG','PRK')

load_gwas <- function(g) {
  base <- file.path(outdir_root, g, 'magma', 'interaction')
  r <- fread(file.path(base, 'interaction_results.tsv'))
  c <- fread(file.path(base, 'interaction_calibration.tsv'))
  # The interaction module writes 17 cols on eff5452+; earlier SCZ output
  # was patched post-hoc to append the two set-level diagnostic columns.
  stopifnot('set_n_tissues_p_low' %in% names(r),
            'set_bias_suspect'    %in% names(r))
  r[, gwas := g]; c[, gwas := g]
  list(results = r, calib = c)
}
all_data <- lapply(gwas_list, load_gwas)
results  <- rbindlist(lapply(all_data, `[[`, 'results'))
calib    <- rbindlist(lapply(all_data, `[[`, 'calib'))
```

## Headline table per GWAS

`n_fdr_05` = rows passing BH-FDR<0.05 within a GWAS across the full
tissue x set grid. `n_retained` = rows passing FDR AND at least one of
the two de-Leeuw-2018 follow-up tests (top-25% or outlier-removed).
`n_bias_suspect_sets` counts distinct gene sets with p_interaction<1e-3
in >=5 tissues within that GWAS — the fingerprint of a set-level effect
leaking across unrelated tissues (see Section
"Interpreting the bias flag" below).

```{r interaction-headline, eval = T}
headline <- results[, .(
  n_rows              = .N,
  n_tests             = sum(!is.na(p_interaction)),
  pct_tested          = round(100 * sum(!is.na(p_interaction)) / .N, 1),
  n_fdr_05            = sum(!is.na(fdr) & fdr < 0.05),
  n_bonf_tissue_05    = sum(!is.na(p_bonferroni_tissue) & p_bonferroni_tissue < 0.05),
  n_retained          = sum(retained, na.rm = TRUE),
  n_bias_suspect_sets = uniqueN(paste(gmt, gene_set)[set_bias_suspect])
), by = gwas]

calib_summary <- calib[, .(
  n_tissues_finite = sum(is.finite(lambda_gc)),
  lambda_min       = round(min(lambda_gc, na.rm = TRUE), 3),
  lambda_median    = round(median(lambda_gc, na.rm = TRUE), 3),
  lambda_max       = round(max(lambda_gc, na.rm = TRUE), 3),
  median_p_median  = round(median(median_p, na.rm = TRUE), 3)
), by = gwas]

headline <- merge(headline, calib_summary, by = 'gwas', sort = FALSE)
setorder(headline, gwas)
knitr::kable(headline, caption = 'MAGMA gene-set x tissue interaction: per-GWAS headline')
```

## Calibration across tissues

Boxplot of lambda_GC across the 54 GTEx v8 tissues per GWAS. Under a
well-calibrated null the lambdas should sit around 1. All 9 GWAS pass
this basic check (lambda medians 0.88-0.98).

```{r interaction-calibration-plot, fig.width=8, fig.height=4.5, eval = T}
ggplot(calib, aes(x = gwas, y = lambda_gc)) +
  geom_hline(yintercept = 1, colour = 'grey60', linetype = 'dashed') +
  geom_boxplot(outlier.size = 0.6, fill = 'grey85') +
  geom_jitter(width = 0.15, alpha = 0.4, size = 0.7) +
  labs(x = NULL, y = expression(lambda[GC]),
       title = 'Per-tissue calibration of MAGMA interaction p-values',
       subtitle = '54 tissues per GWAS') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Hits and bias flags per GWAS

```{r interaction-hits-plot, fig.width=9, fig.height=5, eval = T}
sig_long <- melt(headline[, .(gwas, n_fdr_05, n_bonf_tissue_05,
                              n_retained, n_bias_suspect_sets)],
                 id.vars = 'gwas', variable.name = 'metric',
                 value.name = 'count')
sig_long[, metric := factor(metric,
                            levels = c('n_fdr_05','n_bonf_tissue_05',
                                       'n_retained','n_bias_suspect_sets'),
                            labels = c('BH-FDR<0.05 rows',
                                       'Per-tissue Bonf<0.05 rows',
                                       'Retained rows',
                                       'Bias-suspect sets'))]
ggplot(sig_long, aes(x = gwas, y = count, fill = metric)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  geom_text(aes(label = count), position = position_dodge(width = 0.85),
            vjust = -0.3, size = 3) +
  labs(x = NULL, y = 'count',
       title = 'Interaction hits and bias flags per GWAS',
       fill = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = 'bottom')
```

## Top interaction hits (p_interaction < 1e-4) across GWAS

Rows below list every interaction with a raw p < 1e-4 pooled across the
9 GWAS. `p_top25` and `p_interaction_outliers_removed` are the two
de-Leeuw-2018 follow-up p-values — a real tissue-specific effect
should retain a small p through both. `set_bias_suspect` is the
set-level bias flag from the wrapper.

```{r interaction-top-hits, eval = T}
top_hits <- results[!is.na(p_interaction) & p_interaction < 1e-4,
                    .(gwas, tissue, gene_set, gmt, n_genes_in_set,
                      beta = round(beta, 3),
                      p_interaction = signif(p_interaction, 3),
                      fdr = signif(fdr, 3),
                      p_bonferroni_tissue = signif(p_bonferroni_tissue, 3),
                      p_top25 = signif(p_top25, 3),
                      p_out = signif(p_interaction_outliers_removed, 3),
                      n_outliers,
                      set_bias_suspect,
                      set_n_tissues_p_low,
                      retained)]
setorder(top_hits, p_interaction)
knitr::kable(head(top_hits, 25),
             caption = 'Top 25 gene-set x tissue interaction hits, p<1e-4')
```

## Retained hits (survive both follow-up tests)

The rows most defensible for the paper — pass primary BH-FDR AND at
least one follow-up test at p<0.05, AND are NOT flagged bias-suspect.

```{r interaction-retained, eval = T}
retained_clean <- top_hits[retained == TRUE & set_bias_suspect == FALSE]
knitr::kable(retained_clean,
             caption = 'Retained interaction hits, non-bias-flagged')
```

## Interpreting the bias flag

On the initial genome-wide COAD01 test-drive of this module, several
brain/nervous-system-related gene sets (GOBP_POSTSYNAPSE_ORGANIZATION
and neighbours) surfaced with nominally-significant p_interaction in
20-40 of the 54 tissues — including many peripheral tissues where a
tissue-specific brain-pathway signal is a priori implausible. A manual
top-25% follow-up on POSTSYNAPSE x Breast_Mammary_Tissue confirmed the
bias: primary p=1.65e-6 collapsed to p=0.003, meaning the highest-Breast-
residual genes within the set did NOT concentrate the signal, as they
should under a real tissue-specific effect. The set-level bias
(commit `eff5452`) counts, per set, how many tissues have
p_interaction<1e-3; sets with >=5 are flagged. Under the global null,
this joint event has probability ~0, so any non-zero count is
informative. Sets flagged in this column should be treated with
caution — they typically capture within-set gene heterogeneity
(correlated with a within-set expression axis not captured by
`Average`) rather than a tissue-specific pathway effect.

```{r interaction-bias-sets, eval = T}
bias_sets <- unique(results[set_bias_suspect == TRUE,
                            .(gwas, gmt, gene_set, n_genes_in_set,
                              set_n_tissues_p_low)])
setorder(bias_sets, -set_n_tissues_p_low)
knitr::kable(head(bias_sets, 25),
             caption = 'Top bias-suspect sets across the 9 GWAS')
```

## Restricted to conditionally-independent tissues

de Leeuw et al. 2018 restricted the tissue axis to a small number of
tissues that (a) were marginally significant in the gene-property step
and (b) survived the conditional (independence) analysis. The
interaction module by design deviates from that (all 54 tissues, to
avoid missing tissue-specific pathway effects hiding under a null
marginal tissue association). But it is worth reporting the restricted
analysis alongside: it slashes the multiple-testing burden and focuses
attention on tissues with independent evidence of relevance.

For each GWAS, `magma_tissue_conditional.indep.txt` (from the
`magma_tissue_conditional` rule) lists the conditionally-independent
tissues. Restricting the interaction grid to those tissues and
recomputing BH-FDR within the reduced (tissue x set) grid yields:

```{r interaction-restricted, eval = T}
indep_tissues <- list()
for (g in gwas_list) {
  ind_f <- file.path(outdir_root, g, 'magma',
                     'magma_tissue_conditional.indep.txt')
  indep_tissues[[g]] <- if (file.exists(ind_f)) readLines(ind_f) else character(0)
}

restricted <- rbindlist(lapply(gwas_list, function(g) {
  tis <- indep_tissues[[g]]
  if (length(tis) == 0) return(NULL)
  rr <- results[gwas == g & tissue %in% tis]
  rr[, fdr_restricted        := p.adjust(p_interaction, method = 'BH')]
  rr[, bonf_tissue_restricted := pmin(1, p_interaction * .N), by = tissue]
  rr[, retained_restricted := !is.na(fdr_restricted) & fdr_restricted < 0.05 &
        ((!is.na(p_top25)                        & p_top25 < 0.05) |
         (!is.na(p_interaction_outliers_removed) & p_interaction_outliers_removed < 0.05)) &
        set_bias_suspect == FALSE]
  rr
}), fill = TRUE)

# Grid summary per GWAS
grid_summary <- restricted[, .(
  n_indep_tissues     = uniqueN(tissue),
  n_rows              = .N,
  n_tested            = sum(!is.na(p_interaction)),
  n_fdr_restricted_05 = sum(!is.na(fdr_restricted) & fdr_restricted < 0.05),
  n_retained_restricted = sum(retained_restricted, na.rm = TRUE)
), by = gwas]
knitr::kable(grid_summary,
             caption = 'Restricted (conditionally-independent tissues) interaction analysis: per-GWAS grid summary')

# The surviving hit(s)
knitr::kable(restricted[retained_restricted == TRUE,
                        .(gwas, tissue, gene_set, gmt, n_genes_in_set,
                          beta = round(beta,3),
                          p_interaction = signif(p_interaction,3),
                          fdr_restricted = signif(fdr_restricted,3),
                          fdr_global = signif(fdr,3),
                          p_top25 = signif(p_top25,3),
                          p_out_rm = signif(p_interaction_outliers_removed,3),
                          set_bias_suspect)],
             caption = 'Interactions passing restricted FDR AND at least one follow-up test AND not bias-flagged')
```

Note that hits found in the full 54-tissue analysis but in a tissue
that is NOT in that GWAS's independent-tissue list drop out here. In
particular the ADHD Stomach x endosome global hits are lost because
Stomach is not a conditionally-independent enriched tissue for ADHD
(independent tissues are cerebellar only).

## Interaction results for main-effect FDR-significant gene sets

The interaction analysis is underpowered as a discovery scan across
~150K (set x tissue) cells per GWAS - only 3 rows survive global FDR
(all in ADHD Stomach) and 1 survives the restricted-tissue FDR
(ALZ Whole_Blood x Endocytosis). A more informative use of the same
outputs is as a *follow-up annotation* on gene sets already picked up
by the standard MAGMA pathway analysis (main effect, no interaction):
for each main-effect FDR<0.05 set, we can read off which tissue the
interaction test points to, and whether that tissue is one this GWAS
independently associates with. This slashes the multiple-testing burden
(~54 x n_main_sig_sets instead of the full grid) and lets the
interaction module serve as a *"which tissue is driving this pathway?"*
diagnostic.

Two caveats up front:
1. The interaction module requires `n_genes_in_set >= 100` (MAGMA's
   `interaction-sc-size` hard floor). Many main-effect FDR-sig sets are
   smaller than this and drop out of the join - ALS, PRK and ADHD lose
   their main-sig sets entirely for this reason.
2. Power is still modest. Most `min_int_p` values are 1e-3 to 1e-2,
   not extreme. Read the tissue pointer as *directional* evidence
   rather than as a formal significance test.

```{r interaction-vs-main, eval=T, echo=T}
# Load standard main-effect pathway enrichment per GWAS (already
# produced by the magma_pathway rule for both msigdb GMTs) and pool
# BH-FDR across both gmts within each GWAS - same convention as the
# rest of the report.
load_main <- function(g) {
  f2 <- file.path(outdir_root, g,
                  'magma/pathway/c2.all.v2026.1.Hs.entrez/magma_pathway.gsa.out')
  f5 <- file.path(outdir_root, g,
                  'magma/pathway/c5.go.v2026.1.Hs.entrez/magma_pathway.gsa.out')
  d <- rbind(
    fread(cmd = paste("grep -v '^#'", shQuote(f2)))[, gmt := 'c2.all.v2026.1.Hs.entrez'],
    fread(cmd = paste("grep -v '^#'", shQuote(f5)))[, gmt := 'c5.go.v2026.1.Hs.entrez'],
    fill = TRUE)
  d[, gwas := g]
  nm <- if ('FULL_NAME' %in% names(d)) 'FULL_NAME' else 'VARIABLE'
  d[, gene_set := as.character(get(nm))]
  d[, fdr_main := p.adjust(P, method = 'BH')]
  d
}
main <- rbindlist(lapply(gwas_list, load_main), fill = TRUE)

# Per GWAS: how many main-sig sets, and how many are large enough to have
# been tested in the interaction module?
main_sig <- main[fdr_main < 0.05,
                 .(gwas, gene_set, gmt,
                   n_genes_main = NGENES, beta_main = BETA,
                   p_main = P, fdr_main)]
n_main_sig    <- main_sig[, .(n_main_sig = .N), by = gwas]
in_interaction <- unique(results[, .(gwas, gene_set, gmt)])
n_testable    <- merge(main_sig, in_interaction,
                       by = c('gwas', 'gene_set', 'gmt'))[
                       , .(n_main_sig_testable = .N), by = gwas]
size_summary <- merge(n_main_sig, n_testable, by = 'gwas', all = TRUE)
size_summary[is.na(n_main_sig_testable), n_main_sig_testable := 0L]
setorder(size_summary, gwas)
knitr::kable(size_summary,
             caption = 'Main-effect FDR<0.05 gene sets per GWAS, and how many are large enough to be in the interaction module (n_genes >= 100)')

# Join main-sig sets to their per-tissue interaction rows
joined <- merge(main_sig,
                results[, .(gwas, gene_set, gmt, tissue,
                            n_int = n_genes_in_set,
                            beta_int = beta,
                            p_int = p_interaction,
                            p_top25,
                            p_out = p_interaction_outliers_removed,
                            bias = set_bias_suspect)],
                by = c('gwas','gene_set','gmt'), all.x = TRUE)

# Per set: which tissue is the top interaction hit, is it in the indep set?
per_set <- joined[!is.na(p_int),
                  .(top_int_tissue    = tissue[which.min(p_int)],
                    top_int_p         = min(p_int),
                    top_int_p_bonf54  = pmin(1, min(p_int) * 54),
                    top_indep_tissue  = tissue[
                      tissue %in% indep_tissues[[unique(gwas)]]][
                        which.min(p_int[tissue %in% indep_tissues[[unique(gwas)]]])],
                    top_indep_p       = suppressWarnings(min(p_int[
                                          tissue %in% indep_tissues[[unique(gwas)]]],
                                          na.rm = TRUE)),
                    bias_flagged      = any(bias, na.rm = TRUE)),
                  by = .(gwas, gene_set, gmt, p_main, fdr_main, n_genes_main)]
per_set[is.infinite(top_indep_p), top_indep_p := NA_real_]

# Any main-sig set where the top interaction tissue is an indep tissue at p<0.05
tissue_pointers <- per_set[!is.na(top_indep_p) & top_indep_p < 0.05]
setorder(tissue_pointers, gwas, top_indep_p)

knitr::kable(
  tissue_pointers[, .(gwas,
                      gene_set = substr(gene_set, 1, 55),
                      p_main = signif(p_main, 3),
                      fdr_main = signif(fdr_main, 3),
                      top_int_tissue_global = top_int_tissue,
                      top_int_p_global = signif(top_int_p, 3),
                      top_indep_tissue,
                      top_indep_p = signif(top_indep_p, 3),
                      bias_flagged)],
  caption = 'Main-sig gene sets whose interaction test points at a conditionally-independent tissue (p<0.05)')

# Summary counts
knitr::kable(
  per_set[, .(
    n_main_sig_testable    = .N,
    n_with_bonf54_int_hit  = sum(top_int_p_bonf54 < 0.05, na.rm = TRUE),
    n_pointing_to_indep_lt_05 = sum(top_indep_p < 0.05, na.rm = TRUE),
    n_bias_flagged         = sum(bias_flagged, na.rm = TRUE)
  ), by = gwas],
  caption = 'Per-GWAS summary: how the interaction module annotates main-effect FDR-sig sets')
```

The interpretable pattern that emerges:

- **ALZ** (Whole_Blood + Spleen independent): main-sig sets that also
  show interaction signal are all immune / endocytic - Receptor
  Mediated Endocytosis (Whole_Blood, p=8.8e-6), Adaptive Immune
  Response (Whole_Blood, p=1.7e-4), Positive Regulation of Endocytosis
  (Spleen, p=0.016), Cell Surface (Spleen, p=0.023), Immune Response
  Regulating Cell Surface (Spleen, p=0.019). Coherent with peripheral
  immune / APOE / TREM2 biology.
- **SCZ** (cortex + cerebellum + Pituitary independent): synaptic /
  neuron sets point to Brain Frontal Cortex BA9 (Neuron Projection,
  p=0.002) and Pituitary (Synapse, Cell-Cell Signaling, Synaptic
  Signaling, Trans-Synaptic Signaling - all p in 0.007 to 0.048).
  Pituitary directionality is consistent with SCZ HPA-axis /
  neurohormonal literature.
- **MDD** (cortex + cerebellum + Pituitary independent): synapse
  organisation and neuron-projection sets point to Brain_Frontal_Cortex
  and Brain_Cerebellar_Hemisphere. One CELL_JUNCTION_ASSEMBLY hit is
  bias-flagged and should be treated with caution.
- **BIP, MIG**: no main-sig set produces an interaction hit in an
  independent tissue.
- **ALS, PRK, ADHD, ASD**: no main-sig sets survive the interaction
  module's set-size filter (or, for ADHD/ASD, no main-sig sets at all
  at FDR<0.05), so this annotation is empty for them.

Read as *hypothesis generation* rather than formal discovery: for a
main-effect pathway that's already survived FDR, the interaction test
tells you which of the GWAS-relevant tissues might be driving it. The
ALZ endocytosis-in-blood pointer is the strongest example - it
recovers known biology and is the same signal that survives the strict
restricted-tissue discovery scan.

## QQ plots per GWAS

Thin the mass by showing only the 5,000 smallest p-values per GWAS; the
tail is where any real signal must live.

```{r interaction-qq, fig.width=9, fig.height=8, eval = T}
qq_dt <- results[!is.na(p_interaction)]
qq_dt[, p_rank   := frank(p_interaction, ties.method = 'average'), by = gwas]
qq_dt[, n        := .N, by = gwas]
qq_dt[, expected := -log10(p_rank / (n + 1))]
qq_dt[, observed := -log10(p_interaction)]

ggplot(qq_dt[p_rank <= 5000], aes(x = expected, y = observed)) +
  geom_abline(intercept = 0, slope = 1, colour = 'red', linetype = 'dashed') +
  geom_point(alpha = 0.3, size = 0.5) +
  facet_wrap(~ gwas, ncol = 3, scales = 'free') +
  labs(x = expression(-log[10](p[expected])),
       y = expression(-log[10](p[observed])),
       title = 'QQ plot of interaction p-values (top 5k per GWAS)') +
  theme_bw()
```

