---
title: "GenoDisc Pipeline - Technical Details"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    toc_depth: 2
    css: styles/styles.css
    includes:
      in_header: header.html
      after_body: footer.html

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(knitr)
library(data.table)
```

***

This document describes the methods and tools used by the GenoDisc pipeline. For instructions on how to install and run the pipeline, see the [User Guide](pipeline_guide.html).

***

# Software and dependencies

The pipeline relies on the following tools:

```{r, eval=T, echo=F, results='asis'}
tools <- list(
  list(Tool = 'Snakemake', Purpose = 'Workflow management', Reference = '[Molder et al. 2021](https://doi.org/10.12688/f1000research.29032.2)'),
  list(Tool = 'LDSC', Purpose = 'LD score regression for heritability and genetic correlation', Reference = '[Bulik-Sullivan et al. 2015](https://doi.org/10.1038/ng.3211)'),
  list(Tool = 'PLINK 1.9', Purpose = 'LD clumping and general variant-level operations', Reference = '[Chang et al. 2015](https://doi.org/10.1186/s13742-015-0047-8)'),
  list(Tool = 'GCTA-COJO', Purpose = 'Conditional and joint association analysis', Reference = '[Yang et al. 2012](https://doi.org/10.1038/ng.2213)'),
  list(Tool = 'SuSiE', Purpose = 'Fine-mapping via Sum of Single Effects model', Reference = '[Wang et al. 2020](https://doi.org/10.1111/rssb.12388)'),
  list(Tool = 'MAGMA', Purpose = 'Gene-level and gene set association analysis', Reference = '[de Leeuw et al. 2015](https://doi.org/10.1371/journal.pcbi.1004219)'),
  list(Tool = 'FUSION', Purpose = 'Transcriptome-wide association study', Reference = '[Gusev et al. 2016](https://doi.org/10.1038/ng.3506)'),
  list(Tool = 'SMR', Purpose = 'Summary-data-based Mendelian randomisation', Reference = '[Zhu et al. 2016](https://doi.org/10.1038/ng.3538)'),
  list(Tool = 'TWAS-GSEA', Purpose = 'Gene set enrichment of TWAS results', Reference = '[Pain et al. 2023](https://doi.org/10.1101/2023.07.24.23293084)'),
  list(Tool = 'DrugTargetor', Purpose = 'Drug-gene interaction database for enrichment', Reference = '[Gaspar et al. 2019](https://doi.org/10.1093/bioinformatics/bty742)'),
  list(Tool = 'liftOver', Purpose = 'Genome build coordinate conversion', Reference = '[UCSC Genome Browser](https://genome.ucsc.edu/cgi-bin/hgLiftOver)')
)

df <- do.call(rbind, lapply(tools, as.data.frame, stringsAsFactors = FALSE))
kable(df, format = 'markdown')
```

***

# Workflow overview

The pipeline is implemented as a Snakemake workflow composed of eight rule modules:

1. **`dependencies.smk`** - Downloads and prepares reference data (1000 Genomes, gene annotations, LD scores, TWAS weights, drug-gene databases)
2. **`sumstat_qc.smk`** - Cleans, harmonises, and quality-controls GWAS summary statistics
3. **`finemapping.smk`** - Performs fine-mapping of GWAS loci
4. **`magma.smk`** - Runs gene-level and gene set association analyses
5. **`twas_pwas.smk`** - Performs transcriptome- and proteome-wide association analyses
6. **`smr.smk`** - Runs summary-data-based Mendelian randomisation
7. **`gcsc.smk`** - Performs gene co-regulation-based drug scoring
8. **`report.smk`** - Packages all results into a single `.rds` object

Reference data and tool dependencies are downloaded automatically on the first pipeline run.

***

# Part 1: Descriptive analyses

## Summary statistic QC {#summary-statistic-qc}

GWAS summary statistics undergo a standardised cleaning procedure implemented in `scripts/sumstat_cleaner.R`:

1. **Column name mapping**: The pipeline recognises a wide range of common column name variants and maps them to a standardised set of column names. For example, `PVALUE`, `PVAL`, `P_VALUE`, and `GC_PVALUE` are all mapped to `P`.

2. **Allele harmonisation**: Alleles are harmonised against the 1000 Genomes reference panel. Strand-ambiguous variants are checked against reference allele frequencies and removed if they cannot be resolved.

3. **Genome build detection and liftover**: If the summary statistics do not match the GRCh37/hg19 build, the pipeline attempts automatic coordinate liftover using the UCSC liftOver tool.

4. **Quality filtering**: Variants are filtered based on:
   - Allele frequency (if available): extreme frequency variants are flagged
   - Imputation quality (INFO score, if available)
   - Duplicate removal
   - Multiallelic variant removal

5. **Effect size conversion**: If only odds ratios are provided, they are converted to log-odds (BETA). Z-scores are converted to BETA and SE where possible.

The cleaned summary statistics are stored in `{outdir}/results/{gwas}/gwas_sumstat/`. A munged copy (`.cleaned.munged.sumstats.gz`) is written alongside them, containing `CHR SNP BP A1 A2 Z N` with `Z = BETA/SE`, as required by LDSC and FUSION.

<details><summary>**Recognised column name variants**</summary>

The pipeline accepts many common column name formats. Below are examples of recognised names for key columns:

- **SNP**: SNP, RSID, MARKERNAME, SNPID, ID, VARIANT_ID, HM_RSID, ...
- **CHR**: CHR, CHROMOSOME, CHROM, CHR_BUILD37, CHR_BUILD38, HM_CHROM, ...
- **BP**: BP, POS, POSITION, PHYSPOS, POS_BUILD37, BASE_PAIR_LOCATION, HM_POS, ...
- **A1**: A1, ALLELE1, EFFECT_ALLELE, EA, REF, RISK_ALLELE, HM_EFFECT_ALLELE, ...
- **A2**: A2, ALLELE2, OTHER_ALLELE, NEA, ALT, NON_EFFECT_ALLELE, HM_OTHER_ALLELE, ...
- **BETA**: BETA, EFFECT, B, SIGNED_SUMSTAT, EST, LOG_ODDS, HM_BETA, ...
- **OR**: OR, ODDS_RATIO, ODDS-RATIO, OR(A1), HM_ODDS_RATIO, ...
- **SE**: SE, STDERR, STDER, STANDARD_ERROR, OR_SE, ...
- **P**: P, PVALUE, PVAL, P_VALUE, P-VALUE, WALD_P, P_BOLT_LMM, ...
- **Z**: Z, ZSCORE, Z-SCORE, ZSTAT, GC_ZSCORE, ...
- **N**: N, WEIGHT, TOTALSAMPLESIZE, SAMPLESIZE, N_TOTAL, N_ANALYZED, ...
- **FREQ**: FRQ, FREQ, MAF, AF, EAF, EFFECT_ALLELE_FREQ, HM_EFFECT_ALLELE_FREQUENCY, ...
- **INFO**: INFO, IMPINFO, IMPQUALITY, INFO_UKBB, ...

</details>

***

## LD score regression

[LDSC](https://github.com/bulik/ldsc) (LD Score Regression) is used to estimate:

- **SNP-heritability** ($h^2_{SNP}$): the proportion of phenotypic variance explained by common SNPs
- **LD score regression intercept**: an indicator of confounding or population stratification in the GWAS

The pipeline uses pre-computed LD scores from the European 1000 Genomes reference panel. LDSC is run on the Z-score formatted summary statistics written by `sumstat_cleaner.R` alongside the cleaned sumstats (`.cleaned.munged.sumstats.gz`).

For LDSC compatibility, the cleaner writes a HapMap3-restricted `.cleaned.munged.mergealleles.sumstats.gz` copy (columns `CHR SNP BP A1 A2 Z N`) alongside the full-genome munged file. Only variants present in `w_hm3.snplist` with alleles matching the merge-alleles list are kept; strand-ambiguous SNPs are dropped. Every LDSC step downstream — heritability, rG against a reference panel, and within-list rG — reads this restricted copy so that both sides of any `--rg` call share the same SNP set and allele orientation.

Results are stored in `{outdir}/results/{gwas}/ldsc/`.

***

## Bivariate LDSC (genetic correlation)

Genetic correlation ($r_g$) between GWAS is estimated using bivariate LDSC (`ldsc.py --rg`). Because both sides of each pair are the `.cleaned.munged.mergealleles.sumstats.gz` files produced by the pipeline itself, allele orientation and SNP overlap are guaranteed by construction — the "Incompatible alleles in .sumstats files" error mode is avoided without special handling.

Two independent rG passes are available:

- **Against a reference panel** (`gencor_gwas_list`). One `--rg` call per (primary GWAS × reference trait); the reference list is a `gwas_list`-shaped file whose `path` column points at already-munged `.sumstats.gz` files. Per-pair results and the log per pair are written under `{outdir}/results/{gwas}/gencor/`; the aggregated per-primary CSV (`{gwas}_gencor_res.csv`) applies BH-FDR across every reference trait for that primary.
- **Within the primary list** (`gencor_within_gwas_list: T`). One `--rg` call per unique (primary $i$, primary $j$) pair with $i \ne j$. A helper rule synthesises a per-primary `gencor_within_list.txt` listing every other EUR primary (with `path` pointing at that primary's `.cleaned.munged.mergealleles.sumstats.gz`), then reuses the same shell loop as the reference-panel case. Results land in `{outdir}/results/{gwas}/gencor_within/`. Self-pairs are excluded at list-build time; non-EUR primaries are skipped upstream by the pipeline's standard `gwas_list_df_eur` filter.

Failure handling is identical for both passes: `ldsc.py` exits `0` even on `KeyError: 'Incompatible alleles in .sumstats files'`, so the shell loop writes a `GENCOR_PAIR_FAILED:` marker on non-zero exit and the R post-processor (`scripts/process_ldsc_gencor.R`) treats that marker (and LDSC's own `ERROR computing rg` line) as an `NA` row, keeping successful pairs intact.

***

## LD clumping

LD clumping is performed using PLINK to identify independent genome-wide significant associations. The procedure uses:

- P-value threshold: 5 x 10^-8^
- LD threshold ($r^2$): 0.1
- Clumping window: 250 kb
- Reference panel: 1000 Genomes European samples

Results are stored in `{outdir}/results/{gwas}/clump/`.

***

## GCTA-COJO

[GCTA-COJO](https://yanglab.westlake.edu.cn/software/gcta/) (Conditional and Joint analysis) performs stepwise conditional analysis to identify independently associated variants. Unlike simple LD clumping, COJO accounts for the joint effects of variants using summary statistics and an LD reference panel.

Results are stored in `{outdir}/results/{gwas}/cojo/`.

***

# Part 2: Gene finding analyses

## Fine-mapping (SuSiE)

[SuSiE](https://stephenslab.github.io/susieR/) (Sum of Single Effects) is used for Bayesian fine-mapping. For each genome-wide significant locus, SuSiE identifies credible sets of causal variants. The analysis:

1. Defines loci around genome-wide significant variants
2. Computes LD matrices from the 1000 Genomes European reference panel
3. Fits the SuSiE model allowing for up to L=1 causal variants per locus (configurable)
4. Returns 95% credible sets with posterior inclusion probabilities (PIPs) for each variant

Results are stored in `{outdir}/results/{gwas}/finemap/`.

***

## MAGMA gene-level association

[MAGMA](https://ctg.cncr.nl/software/magma) performs gene-level association analysis by aggregating SNP-level association statistics within gene boundaries. The pipeline:

1. Maps SNPs to genes using Ensembl gene annotations (with a configurable window)
2. Computes gene-level test statistics accounting for LD between SNPs
3. Provides gene-level p-values and effect sizes

Results are stored in `{outdir}/results/{gwas}/magma/`.

***

## Transcriptome-wide association (TWAS)

[FUSION](http://gusevlab.org/projects/fusion/) performs transcriptome-wide association studies by testing whether genetically predicted gene expression is associated with the GWAS phenotype. The pipeline supports multiple eQTL reference panels:

- **GTEx panels**: Tissue-specific eQTL weights from the GTEx consortium (user-specified tissues)
- **PsychENCODE**: Brain-specific eQTL weights from the PsychENCODE consortium
- **Non-GTEx panels**: Additional eQTL panels (e.g. CMC brain RNA-seq)
- **External panels**: User-provided TWAS weight panels in FUSION format

For each panel, the pipeline:

1. Runs TWAS per chromosome
2. Combines results across chromosomes
3. Identifies genome-wide significant gene-trait associations (Bonferroni-corrected)
4. Performs conditional analysis on significant hits to identify independently associated genes

Results are stored in `{outdir}/results/{gwas}/twas/`.

***

## Summary-data-based Mendelian randomisation (SMR)

[SMR](https://yanglab.westlake.edu.cn/software/smr/) tests for causal or pleiotropic associations between gene expression and complex traits using summary-level data from GWAS and eQTL studies. The pipeline supports multiple eQTL/pQTL reference panels:

- **PsychENCODE**: Brain eQTL data
- **MetaBrain**: eQTL data from multiple brain regions (cortex, basal ganglia, cerebellum, hippocampus, spinal cord)
- **eQTLGen**: Blood eQTL data
- **ROSMAP**: Brain pQTL data

The HEIDI (Heterogeneity in Dependent Instruments) test is used to distinguish causal/pleiotropic associations from linkage.

Results are stored in `{outdir}/results/{gwas}/smr/`.

***

## Proteome-wide association (PWAS)

PWAS extends the TWAS framework to protein levels, using pQTL weights to predict genetically determined protein expression. The pipeline supports:

- **ROSMAP**: Brain pQTL weights (n=376)
- **Banner**: Brain pQTL weights (n=152)

PWAS uses the same FUSION framework as TWAS, with protein expression prediction models in place of gene expression models.

Results are stored in `{outdir}/results/{gwas}/pwas/`.

***

# Part 3: Drug finding analyses

## MAGMA DrugTargetor enrichment

MAGMA gene set enrichment analysis is performed using [DrugTargetor](https://drugtargetor.com/) drug-gene interaction sets. This tests whether genes targeted by specific drugs are enriched for GWAS associations. The analysis also evaluates enrichment at the ATC (Anatomical Therapeutic Chemical) classification level.

Results are stored in `{outdir}/results/{gwas}/magma/`.

***

## TWAS-GSEA DrugTargetor enrichment

[TWAS-GSEA](https://github.com/opain/TWAS-GSEA) performs gene set enrichment analysis using TWAS results rather than SNP-level summary statistics. By leveraging gene expression prediction, TWAS-GSEA can detect drug-gene set enrichments that are not captured by traditional MAGMA-based approaches.

Both directional and non-directional enrichment tests are available:

- **Directional**: Tests whether disease-associated genes have expression changes in the same direction as drug effects
- **Non-directional**: Tests for enrichment regardless of effect direction

Results are aggregated across TWAS panels and stored alongside other enrichment results.

***

## Pathway enrichment (MAGMA + TWAS-GSEA)

When `pathway_gmt_dir` is set and `magma_pathway: T` and/or `twas_gsea_pathway: T` are enabled, the same MAGMA gene-set-analysis and non-directional TWAS-GSEA machinery is applied to user-supplied `.gmt` files (typically MSigDB collections). The pipeline globs `*.gmt` in `pathway_gmt_dir` at parse time and fans out one job per (primary GWAS × gmt) for MAGMA, and one per (primary GWAS × TWAS panel × gmt) for TWAS-GSEA.

**Entrez ↔ Symbol conversion.** MSigDB `.gmt` files are typically keyed on Entrez IDs. MAGMA operates directly on Entrez IDs (matching the `NCBI37.3.gene.loc` gene table it already uses), so no conversion is needed on that side. TWAS-GSEA matches gene set members to gene symbols instead, so a small pre-processing rule (`scripts/convert_gmt_entrez_to_symbol.R`) rewrites each `.gmt` into a symbol-keyed copy under `{outdir}/results/pathway_gmts/` using the same `NCBI37.3.gene.loc` file as the mapping. Genes without a symbol match are dropped with a per-gmt count logged; the user's input directory is never modified.

**Statistics.** Both sides mirror the DrugTargetor equivalents:

- MAGMA runs a competitive gene-set test per gene set, emitting `NGENES, BETA, SE, P` in `magma_pathway.gsa.out`.
- TWAS-GSEA runs its non-directional right-tail enrichment (probit link on $1 - P_{\text{TWAS}}$) per (panel, gmt) pair, emitting `Estimate, SE, T, P` in `.competitive.txt`.

Per-gmt cleaned CSVs land under `{outdir}/results/{gwas}/magma/pathway/{gmt}/` and `{outdir}/results/{gwas}/twas/pathway/{gmt}/` with per-set BH-FDR **deliberately deferred** to bundle-write time.

**Pooled FDR.** The `read_pathway_magma()` and `read_pathway_twas_gsea()` helpers in `package_results_functions.R` `rbind` every per-gmt CSV for a primary before applying `p.adjust(P, method = "BH")` once — for TWAS-GSEA the pool also includes every TWAS panel. This means the reported `P.FDR` correctly reflects the full multiple-testing burden of the pathway pass, not the per-file burden.

***

## TWAS-GSEA CMAP drug signatures

In addition to DrugTargetor gene sets, TWAS-GSEA is also applied using reprocessed [CMAP](https://clue.io/) (Connectivity Map) level 5 drug expression signatures. This identifies drugs that reverse the disease-associated gene expression pattern, which is a key strategy in computational drug repurposing.

The CMAP analysis tests each drug's expression signature against the TWAS results to identify compounds whose effects on gene expression oppose the disease-associated pattern. Results include enrichment statistics and drug annotations including mechanism of action (MOA).

***

## Gene co-regulation scoring (GCSC)

GCSC (Gene Co-regulation-based Scoring) uses tissue-specific gene co-regulation networks from GTEx to score drug-gene interactions. Rather than testing individual gene associations, GCSC evaluates whether the co-regulation structure of drug target genes is associated with the GWAS phenotype.

The analysis:

1. Runs TWAS using GTEx v7 weights for specified tissues
2. Prepares drug-gene interaction sets from DrugTargetor
3. Computes co-regulation-based enrichment scores for each drug

Results include drug-level scores and ATC classification enrichment, stored in `{outdir}/results/{gwas}/gcsc/`.

***

## MAGMA tissue enrichment

MAGMA tissue enrichment analysis tests whether genes expressed in specific tissues are enriched for GWAS associations. This uses GTEx tissue expression profiles to identify tissues where GWAS-associated genes are preferentially expressed, and performs conditional analysis to identify independently enriched tissues.

**Conditional analysis.** The marginal test evaluates each GTEx tissue independently, so several tissues can appear enriched simply because their expression profiles overlap with the true driving tissue rather than carrying independent signal. To resolve this, `scripts/magma_tissue_conditional.R` takes the tissues that are FDR-significant (`P.FDR ≤ 0.05`) in the marginal test, orders them by p-value, and iteratively re-tests each remaining tissue while conditioning on the current lead tissue's expression profile (via MAGMA's `condition-hide`, alongside the `Average` expression covariate).

A tissue is **retained** if any of the following hold after conditioning:

- **Collinear with the lead** — the conditional model cannot separate its signal from the lead's (variance-explained is undefined / `NA`). The tissue is kept because it is statistically indistinguishable from the lead, not because it adds independent information.
- **Comparable marginal strength** — its marginal χ² is within 95% of the lead's (`R2_ratio > 0.95`), i.e. the two tissues are essentially co-leading.
- **Still significant after conditioning** — its conditional `P < 0.05`, meaning it contributes signal beyond what the lead explains.

Tissues that fail all three criteria are dropped, because their marginal enrichment is adequately accounted for by tissues already retained. The final set is written to `magma_tissue_conditional.indep.txt`. A retained tissue should be interpreted as a plausible driver of the GWAS signal, but retention does not imply it acts independently of every other retained tissue: some are kept precisely because collinearity prevents that distinction.

Results are stored in `{outdir}/results/{gwas}/magma/`.

***

# Results packaging

The `package_results` rule (`rules/report.smk`) aggregates every analysis into a split-per-block bundle:

- `{outdir}/results/package/` — one `.rds` per analysis block (`gwas_qc`, `snp_assoc`, `mol_assoc/exp/fusion`, `tx/drug`, `tx/pathway`, `tissue`, …) organised as `gwas/<GWAS>/<block>.rds`, plus `configuration/config.rds`, `configuration/gwas_list.rds`, `manifest.json`, and a copy of `reader.R` so the bundle is self-contained.
- `{outdir}/results/bundle.tar.gz` — the tarballed `package/` directory. This is the file the Shiny explorer accepts and the download the web app hands users.

The canonical block IDs are defined in `scripts/reader.R:.gd_block_ids`. Downstream code loads blocks lazily via `gd_open()` → `gd_read(gd, "<gwas>", "<block>")` so that memory usage scales with the block being inspected rather than with the full bundle — see the [User Guide](pipeline_guide.html#reading-results-into-r) for examples.

Pooled-FDR helpers in `scripts/functions/package_results_functions.R` (e.g. `read_pathway_magma()`, `read_twas_gsea_drug()`, `read_twas_gsea_cmap_drug()`) compute BH-adjusted p-values across the natural pooling unit (all gmts for pathway, all panels for TWAS-GSEA, all CMAP signatures × panels for CMAP) at bundle-write time, so the FDR column in each block reflects the correct multiple-testing burden without the consumer having to re-adjust.

For backwards compatibility, `gd_open()` also accepts a legacy monolithic `results_package.rds` produced by older pipeline runs.

***

For questions or issues, please contact Oliver Pain (oliver.pain@kcl.ac.uk) or open an issue on the [GitHub repository](https://github.com/opain/GenoDisc).
